---
title: Empirical laws, old and new, in the nutrition transition
author:
  - name: Benjamin Schiek
    email: b.schiek@cgiar.org
    affiliation: Alliance Bioversity-CIAT
    footnote: Corresponding Author
address:
  - code: Alliance Bioversity-CIAT
    address: Alliance Bioversity-CIAT, Km 17 Recta Cali-Palmira, Valle del Cauca, Colombia
abstract: |
  Engel's Law and Bennett's Law are often mentioned in the nutrition transition literature. But seldom is any evidence of their validity presented in a straightforward, visual manner. Here I present...using FAO and World Bank data. I also present variant formulations which fit the data more cleanly. Including consumption inequality in the model improves the fit. The fitted slope and y-intercept parameters are fairly stable across four and a half decades of data. Animations (available as supplemental material) offer striking visual confirmation of this stability. I propose a new theoretical framework that unifies the two laws as separate manifestations of a single underlying principle. Finally, I present another empirical law that appears to have gone unnoticed until now: a tight log-linear corresponance between commodity expected price and price volatility.
keywords: Nutrition transition, Bennett's Law, Engel's Law, Global food demand, Food price volatility
journal: "CG Space"
bibliography: Empirical laws, old and new, in the nutrition transition.bib
header-includes:
  - \usepackage{float}
  - \floatplacement{figure}{H}
#linenumbers: true
numbersections: true
csl: elsevier-harvard.csl
output:
    rticles::elsevier_article:
    latex_engine: xelatex
mainfont: Garamond
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      message = FALSE,
                      warning = FALSE)
library(plyr)
library(tidyverse)
library(patchwork)
library(ggrepel)
#library(kableExtra)
shape_vec <- c(21:24, 4)
point_size <- 2
label_size <- 4

#==========================================================================
shift_legend2 <- function(p) {
  # ...
  # to grob
  gp <- ggplotGrob(p)
  facet.panels <- grep("^panel", gp[["layout"]][["name"]])
  empty.facet.panels <- sapply(facet.panels, function(i) "zeroGrob" %in% class(gp[["grobs"]][[i]]))
  empty.facet.panels <- facet.panels[empty.facet.panels]
  
  # establish name of empty panels
  empty.facet.panels <- gp[["layout"]][empty.facet.panels, ]
  names <- empty.facet.panels$name
  # example of names:
  #[1] "panel-3-2" "panel-3-3"
  
  # now we just need a simple call to reposition the legend
  #lemon::reposition_legend(p, 'center', panel=names)
  p_out <- lemon::reposition_legend(p, 'center', panel=names)
  #class(lemon::reposition_legend(p, 'center', panel=names))
  return(p_out)
}
#==========================================================================

```

# Introduction

A good understanding of Bennett's Law [@bennett1941international] and Engel's Law [-@zimmerman1932ernst] is key to understanding the future composition and quantity of food demand, at all scales. The laws figured centrally in the food crisis literature of the 1960s [@brown1963man]; and, nowadays, form the _implicit_ premise of much of the Nutrition Transition literature []. However, this time around there is surprisingly little explicit reference to or graphical representation of the laws. It appears, moreover, that many aid bureaucrats, and even some researchers, are simply unaware of their existence. Tilman et al. [-@tilman2011global], for example, present what they believe to be a newly discovered empirical relation between Gross Domestic Product (GDP) and calorie consumption; whereas this is probably better characterized as a corrollary of the empirical relation observed by Engel in 1841. International breeding programs, meanwhile, continue to emphasize cereals over non-starchy crops well in excess of what would be justified by Bennett's Law [EIB report].

Here, I open a parenthesis in the Nutrition Transition discussion to 
For these reasons, it is worhtwhile to open a parenthesis in the Nutrition Transition discussion in order to take stock of and appreciate the relevance of these empirical laws in policy making contexts. Here, I make a contribution in this direction by presenting a graphical analysis of Bennett's and Engel's Laws in greater detail than has so far appeared in the literature. I also explore variant formulations of the laws that improve the fit with the data. Animations of the analysis^[Presented in the supplemental materials.] provide striking confirmation of the stability of these laws over time. I then discuss briefly how Bennett's and Engel's Laws can be thought of as two manifestations of a single mathematical law. Finally, I present a new empirical law that appears to govern the relation between a commodity's expected price and its price risk. This new law may serve to open up new perspectives on the role of price risk in food system resilience and nutritional security---a topic which has so far received little attention.
<!-- I then propose a theoretical unification of the two laws by deriving them both from the same principle of diminishing returns to marginal increases in expenditures on inferior goods. -->

# Bennett's Law

>"The broad fact that variety rather than uniformity is characteristic of food consumption throughout the world is familiar and indisputable. Differences are found in the 'dishes' consumed from country to country, region to region, locality to locality, even family to family. Within a country or region, these differences appear also from one epoch to another. They are so prevalent and conspicuous, in fact, that it seems almost hopeless to discover, in the international sphere, some way of characterizing summarily the diets of different nations and contrasting them one with the other" [@bennett1941international].

Even within a fairly small geographical region of Africa, diets can vary considerably (Figure \ref(fig:dietDiversity)). If we were to look at each of these food catgories in more detail, we would see even more diversity in terms of what specific crops constitute the groups.

```{r dietDiversity, fig.show = "hold", fig.width = 7, fig.height=4, fig.align="center", fig.cap="Diets are diverse.", echo = FALSE}
this_file <- "Bennetts Law post_diverse diets.csv"
df_plot <- read.csv(this_file, stringsAsFactors = F)
df_plot$X.1 <- NULL
df_plot$X <- NULL
df_plot <- df_plot[, c("Area", "Year", "Item", "kcal.share", "mu_food_area_item")]
colnames(df_plot)[4] <- "Share of daily diet"
df_plot$Item[grep("Sugar", df_plot$Item)] <- "Sugar\n& sweeteners"
df_plot$Item[grep("Animal", df_plot$Item)] <- "Animal\nproducts"
df_plot$Item[grep("Roots", df_plot$Item)] <- "Starchy\nroots"
#-----------------------------------------------------------------------------
area_vec <- c("Kenya", "Uganda", "Ethiopia", "Tanzania", "Rwanda")
df_plot <- subset(df_plot, Area %in% area_vec &
                    Year > 1992)
df_plot$Year <- as.integer(df_plot$Year)
yr_min <- min(df_plot$Year)
yr_max <- max(df_plot$Year)
n <- length(unique(df_plot$Item))
bag_of_colors <- randomcoloR::distinctColorPalette(k = 2 * n)
color_vec <- sample(bag_of_colors, n)
#good_colors <- c("#B346E2", "#74B4D0", "#E16953", "#D6E1D8", "#C7E84D", "#6BE47C", "#E4B64D")
#color_vec <- good_colors
df_plot <- df_plot %>% group_by(Area, Item) %>%
  mutate(mu_food_area_item = mean(`Share of daily diet`, na.rm = T)) %>%
  as.data.frame()
df_plot$Item <- factor(df_plot$Item,
                       levels = unique(df_plot$Item[order(df_plot$mu_food_area_item, df_plot$Item, decreasing = T)]),
                       ordered = T)
gg <- ggplot(df_plot, aes(x = Year, y = `Share of daily diet`, fill = Item))
gg <- gg + geom_area(position = "stack")
gg <- gg + scale_x_continuous(breaks = seq(yr_min, yr_max, length.out = 4))
gg <- gg + scale_fill_manual(values = color_vec)
gg <- gg + facet_wrap(~Area, ncol = 3)
gg <- gg + labs(caption = "Source: FAO")
gg <- gg + theme(legend.title = element_blank(),
                 axis.title.x = element_blank())
gg <- gg + guides(fill = guide_legend(nrow = 4, byrow = T))
gg <- shift_legend2(gg)
gg <- ggplotify::as.ggplot(gg)


```
Nonetheless, Bennett found the "way of characterizing summarily the diets of different nations" he was looking for by examining the carbohydrate content of diets in relation to income. An example is given in the left panel of Figure \ref{fig:Bennet01}. The relation says that, regardless of geography and culture, carbohydrates form a declining portion of the diet as income rises.

A question remains as to how much of the carbohydrates are replaced by proteins and how much are replaced by fats, as income rises. Cursory exploration of the data reveals that the carbohydrates are replaced almost exclusively by fats (right panel, Figure \ref{fig:Bennet01}). This means that the protein share of diets remains essentially constant at all income levels. Metabolically speaking, constant protein intake in the face of declining carbohydrate intake implies increased consumption of animal products. (The increased fat intake also implies this to some degree, although not necessarily since fats may be acquired from non-animal sources.) The often mentioned dietary transition to animal products that accompanies rising incomes and urbanization may thus be considered a corrollary of Bennett's Law.

```{r Bennett01, fig.show = "hold", fig.width = 6, fig.height=5, fig.align="center", fig.cap="\\italic{(Left)} Bennett's Law, \\italic{(Right)} the fat-carb frontier, FAO data for 2017. The size of the points in the fat carb frontier correspond to the log of GDP.", echo = FALSE}
# 4 calories in a gram of protein or carbohydrate, 9 calories in a gram of fat.
# https://www.nal.usda.gov/fnic/how-many-calories-are-one-gram-fat-carbohydrate-or-protein
#===========================================================================
this_file <- "Bennetts Law data.csv"
df_bennett <- read.csv(this_file, stringsAsFactors = F)
#colnames(df_bennett)
df_bennett$X <- NULL
#colnames(df_bennett) <- gsub(".", "\\", colnames(df_bennett), fixed = T)
colnames(df_bennett)[3:9] <- c("Food supply (kcal/capita/day)",
                               "Fat supply quantity (kcal/capita/day)",
                               "Protein supply quantity (kcal/capita/day)",
                               "Carb supply quantity (kcal/capita/day)",
                               "Fat share of diet (%)",
                               "Carb share of diet (%)",
                               "Carb share of diet (%), logged")
colnames(df_bennett)[11] <- "GDP / capita (USD), logged"
#===========================================================================
# Bennett's Law
df_plot <- subset(df_bennett, Year == "2017")
df_mod <- df_plot[, c("Carb share of diet (%), logged", "GDP / capita (USD), logged")]
mod <- lm(`Carb share of diet (%), logged` ~., data = df_mod)
#summary(mod)
m <- round(mod$coefficients[2], 2)
b <- round(mod$coefficients[1], 2)
#-----------------------------------------------------------------------------
label_Africa <- c("Tanzania", "Uganda", "Ethiopia", "South Africa",
                  "Ghana")
label_NAfricaWAsia <- c("Turkey", "Saudi Arabia")
label_Asia <- c("India", "Iran", "China, mainland", "Vietnam", "Japan",
                "South Korea")
label_NAmEurAusNZ <- c("Italy", "France", "Australia")
label_LAC <- c("Nicaragua", "Colombia", "Brazil")
labelThese_vec <- c(label_Africa, label_NAfricaWAsia, label_Asia,
                    label_NAmEurAusNZ, label_LAC)
#-----------------------------------------------------------------------------

df_plot$label_these <- NA
u <- df_plot$Country
df_plot$label_these[which(u %in% labelThese_vec)] <- df_plot$Country[which(u %in% labelThese_vec)]

n <- length(unique(df_plot$Region))
bag_of_colors <- randomcoloR::distinctColorPalette(k = 2 * n)
color_vec <- sample(bag_of_colors, n)

gg <- ggplot(df_plot, aes(x = `GDP / capita (USD), logged`,
                          y = `Carb share of diet (%), logged`,             
                          group = Region, fill = Region,
                          shape = Region,
                          label = label_these))
# gg <- gg + geom_point(aes(fill = Region, group = Region, shape = Region), pch = 21, alpha = 0.5, size = 6, color = "black")
gg <- gg + geom_smooth(aes(group = NULL, fill = NULL, shape = NULL), method = lm, se = F)
gg <- gg + geom_point(alpha = 0.6, size = point_size, color = "black", stroke = 1)
gg <- gg + scale_fill_manual(values = color_vec)
gg <- gg + scale_shape_manual(values = shape_vec)
# gg <- gg + geom_abline(slope = m, intercept = b,
#                        linetype = "dashed", size = 1)
gg <- gg + geom_text_repel(color = "black", size = label_size)
gg <- gg + labs(#title = "Engel's Law", 
  subtitle = paste0("Slope = ", m, ", y-intercept = ", b))#,
#caption = "Source: Author's creation based on World Bank International Comparison Program Data for 2017.
#GDP/capita is based on 2015 prices.")
gg <- gg + guides(fill = guide_legend(nrow = 2, byrow = T, override.aes = list(linetype = 0)),
                  color = guide_legend(override.aes = list(linetype = 0)))
gg <- gg + theme(legend.position = "bottom",
                  legend.spacing.x = unit(0.25, 'cm'),
                  legend.title = element_blank())
gg_BennettsLaw <- gg
#=========================================================================
# Fat-Carb Frontier
df_mod <- df_plot[, c("Carb share of diet (%)", "Fat share of diet (%)")]
mod <- lm(`Carb share of diet (%)` ~., data = df_mod)
#summary(mod)
m <- round(mod$coefficients[2], 2)
b <- round(mod$coefficients[1], 2)

# df_plot$Country[which(df_plot$`Fat share of diet (%)` ==
#                         max(df_plot$`Fat share of diet (%)`))]
df_plot <- subset(df_plot, Country != "Slovakia")
#colnames(df_plot)
df_plot$`GDP / capita (USD)` <- exp(df_plot$`GDP / capita (USD), logged`)
gg <- ggplot(df_plot, aes(x = `Fat share of diet (%)`,
                          y = `Carb share of diet (%)`,
                          group = Region, fill = Region,
                          shape = Region, size = `GDP / capita (USD)`,
                          label = label_these))
gg <- gg + geom_smooth(aes(group = NULL, fill = NULL, shape = NULL), method = lm, se = F)
gg <- gg + geom_point(alpha = 0.6, color = "black", stroke = 1)
gg <- gg + scale_fill_manual(values = color_vec)
gg <- gg + scale_shape_manual(values = shape_vec)
# gg <- gg + geom_abline(slope = m, intercept = b,
#                        linetype = "dashed", size = 1)
gg <- gg + geom_text_repel(color = "black", size = label_size)
gg <- gg + labs(#title = "Bennett's Law", 
  subtitle = paste0("Slope = ", m, ", y-intercept = ", b))#,
#caption = "Source: Author's creation based on World Bank International Comparison Program Data for 2017.
#GDP/capita is based on 2015 prices.")
gg <- gg + theme(legend.position = "bottom",
                legend.spacing.x = unit(0.25, 'cm'),
                legend.title = element_blank())
gg <- gg + guides(fill = guide_legend(nrow = 2, byrow = T, override.aes = list(linetype = 0)),
                  color = guide_legend(override.aes = list(linetype = 0)),
                  size = F)
gg_fatCarbFrontier <- gg
# gg1 + gg2 + plot_layout(ncol = 2, guides = "collect") & theme(legend.position = 'bottom')
#===========================================================================
wrap_plots(list(gg_BennettsLaw, gg_fatCarbFrontier), ncol = 2)

```

The slopes and y-intercepts of both Bennett's Law and the fat-carb frontier are remarkably stable over time (Figure ). An animation of the fat-carb frontier plot is provided in the supplemental materials.

```{r BennettStability, fig.show = "hold", fig.width = 6, fig.height=5, fig.align="center", fig.cap="Parameter stability for \\italic{(Left)} Bennett's Law and \\italic{(Right)} the fat-carb frontier.", echo = FALSE}






```

... Geographically disaggregated...a little less stable.

...estimate future carb demand based on Bennett's Law, compare to actual breeding emphasis on cereals.

# Engel's Law

Engel's Law...Zimmerman [-@zimmerman1932ernst] cites several 19th century datasets that appear to contradict Engel's Law.

The influence of income distribution on Engel's Law was explored by Cirera and Masset [-@cirera2010income]. Following in this vein, here I tried including the income Gini coefficent in the models. is not significant. However, the consumption Gini coefficient is significant at the 0.05 level. Adding consumption inequality to the model increases the adjusted R-squared by 0.10. GCIP data.

Tilman et al. [-@tilman2011global] reached an estimate of 100%-110% increase in food demand based on about 2.5% growth in GDP. More realistic growth estimates ...[@diaz2016future].

"The poorer is a family, the greater is the proportion of the total outgo which must be used for food" [@zimmerman1932ernst].

Only recently African countries included in the ICP data...in Figure \ref{fig:Engel0}

World Bank International Comparison Program Data (ICP)

The influence of income distribution on Engel's Law was explored by Cirera and Masset [-@cirera2010income]. Following in this vein, here I tried including the income Gini coefficent in the models. is not significant. However, the consumption Gini coefficient is significant at the 0.05 level. Adding consumption inequality to the model increases the adjusted R-squared by 0.10. GCIP data.

Tilman et al. [-@tilman2011global] reached an estimate of 100%-110% increase in food demand based on about 2.5% growth in GDP.

"The poorer is a family, the greater is the proportion of the total outgo which must be used for food" [@zimmerman1932ernst].

Only recently African countries included in the ICP data...in Figure \ref(fig:Engel0)

World Bank International Comparison Program Data (ICP)

```{r Engel0, fig.show = "hold", fig.width = 6, fig.height=5, fig.align="center", fig.cap="Engel's Law. Source: Author's creation based on ICP data for 2017. GDP/capita is retrieved from the FAO database and is based on 2015 prices.", echo = FALSE}
#===========================================================================
this_file <- "Engels Law 1.csv"
df_plot <- read.csv(this_file, stringsAsFactors = F)
df_plot$X <- NULL
colnames(df_plot)[3:4] <- c("GDP / capita (USD), logged",
                            "Food expenditure share of GDP, logged")
#---------------------------------------------------------------------------
mod <- lm(`Food expenditure share of GDP, logged` ~ `GDP / capita (USD), logged`, df_plot)
#summary(mod)
m <- round(mod$coefficients[2], 2)
b <- round(mod$coefficients[1], 2)

# unique(df_plot$Country[which(df_plot$Region == "Asia\n(excluding W. Asia)")])
# unique(df_plot$Country[which(df_plot$Region == "N. Africa / W. Asia")])
# label_Africa <- c("Tanzania", "Uganda", "Kenya", "Ethiopia", "South Africa",
#                   "DRC", "Angola", "Ghana","Rwanda", "Mozambique")
# label_NAfricaWAsia <- c("Egypt", "Turkey", "Sudan", "Saudi Arabia", "Jordan",
#                         "Israel", "United Arab Emirates", "Kuwait")
# label_Asia <- c("India", "Iran", "China, mainland", "Vietnam", "Japan",
#                 "South Korea", "Afghanistan", "Indonesia", "Kazakhstan",
#                 "Singapore", "Tajikistan")
# label_NAmEurAusNZ <- c("Austria", "United Kingdom", "France", "Australia",
#                        "United States", "Spain", "Italy")
# label_LAC <- c("Nicaragua", "Haiti", "Argentina", "Colombia", "Mexico",
#                "Dominican Republic", "Brazil", "Uruguay")
# labelThese_vec <- c(label_Africa, label_NAfricaWAsia, label_Asia,
#                     label_NAmEurAusNZ, label_LAC)
label_Africa <- c("Tanzania", "Uganda", "Ethiopia", "South Africa",
                  "Ghana")
label_NAfricaWAsia <- c("Turkey", "Saudi Arabia")
label_Asia <- c("India", "Iran", "China, mainland", "Vietnam", "Japan",
                "South Korea")
label_NAmEurAusNZ <- c("Italy", "France", "Australia")
label_LAC <- c("Nicaragua", "Colombia", "Brazil")
labelThese_vec <- c(label_Africa, label_NAfricaWAsia, label_Asia,
                    label_NAmEurAusNZ, label_LAC)

df_plot$label_these <- NA
u <- df_plot$Country
df_plot$label_these[which(u %in% labelThese_vec)] <- u[which(u %in% labelThese_vec)]

n <- length(unique(df_plot$Region))
bag_of_colors <- randomcoloR::distinctColorPalette(k = 2 * n)
color_vec <- sample(bag_of_colors, n)

shape_vec <- c(21:24, 4)
point_size <- 2
label_size <- 4

gg <- ggplot(df_plot, aes(x = `GDP / capita (USD), logged`,
                          y = `Food expenditure share of GDP, logged`,
                          group = Region, fill = Region, shape = Region,
                          label = label_these))
# gg <- gg + geom_point(aes(fill = Region, group = Region, shape = Region), pch = 21, alpha = 0.5, size = 6, color = "black")
gg <- gg + geom_point(alpha = 0.6, size = point_size, color = "black", stroke = 1)
gg <- gg + scale_fill_manual(values = color_vec)
gg <- gg + scale_shape_manual(values = shape_vec)
# gg <- gg + geom_abline(slope = m, intercept = b,
#                        linetype = "dashed", size = 1)
gg <- gg + geom_smooth(aes(group = NULL, fill = NULL, shape = NULL), method = lm, se = F)
gg <- gg + geom_text_repel(color = "black", size = label_size)
gg <- gg + labs(#title = "Engel's Law", 
  subtitle = paste0("Slope = ", m, ", y-intercept = ", b))#,
#caption = "Source: Author's creation based on World Bank International Comparison Program Data for 2017.
#GDP/capita is based on 2015 prices.")
gg <- gg + guides(fill = guide_legend(nrow = 2, byrow = T, override.aes = list(linetype = 0)),
                  color = guide_legend(override.aes = list(linetype = 0)))
gg <- gg + theme(legend.position = "bottom",
                 legend.spacing.x = unit(0.5, 'cm'),
                 legend.title = element_blank())
gg
#ggsave("Engels Law 1.png", width = 8.5, height = 6.5, units = "in")
```

Better fit when examined by regions except for Sub-Saharan Africa... Note how parameter values vary somewhat by region in Figure \ref(fig:Engel0Disagg). Might also try disaggregating by quantile.

```{r Engel0Disagg, fig.show = "hold", fig.width = 7, fig.height=5, fig.align="center", fig.cap="Engel's Law by region, ICP data for 2017.", echo = FALSE}
region_vec <- unique(df_plot$Region)
facet_labels <- c()
for(i in 1:length(region_vec)){
  this_region <- region_vec[i]
  this_df_plot <- subset(df_plot, Region == this_region)
  mod <- lm(`Food expenditure share of GDP, logged` ~ `GDP / capita (USD), logged`, this_df_plot)
  #summary(mod)
  m <- round(mod$coefficients[2], 2)
  b <- round(mod$coefficients[1], 2)
  #   print(this_region)
  # print(m)
  facet_labels[i] <- paste0("Slope = ", m, ", y-intercept = ", b)
}

names(facet_labels) <- region_vec

gg <- ggplot(df_plot, aes(x = `GDP / capita (USD), logged`,
                          y = `Food expenditure share of GDP, logged`,
                          group = Region, fill = Region, shape = Region,
                          label = label_these))
gg <- gg + geom_point(alpha = 0.6, size = point_size, color = "black", stroke = 1)
gg <- gg + scale_fill_manual(values = color_vec)
gg <- gg + scale_shape_manual(values = shape_vec)
gg <- gg + geom_smooth(method = lm, se = F)
gg <- gg + facet_wrap(~Region, ncol = 3, scales = "free", labeller = labeller(Region = facet_labels))
gg <- gg + geom_text_repel(color = "black", size = label_size)
gg <- gg + theme(strip.background = element_blank(),
                 strip.text.x = element_text(hjust = -0.01),
                 #legend.direction = "vertical",
                 #legend.spacing.y = unit(1, 'cm'),
                 legend.title = element_blank())
gg <- gg + guides(fill = guide_legend(#nrow = 2,
  override.aes = list(linetype = 0)),
  color = guide_legend(override.aes = list(linetype = 0)))
gg <- shift_legend2(gg)
gg <- ggplotify::as.ggplot(gg)
#gg
#ggsave("Engels Law 1 Disaggregated.png", width = 8.5, height = 6.5, units = "in")
```


"Engel's law establishes that as income increases, households' demand for
food increases less than proportionally" [@cirera2010income]... Figure \ref{fig:Engel1}.


```{r Engel1, fig.show = "hold", fig.width = 6, fig.height=5, fig.align="center", fig.cap="Engel's Law, variation 1, ICP data for 2017.", echo = FALSE}
rm(df_plot)
this_file <- "Engels Law 2.csv"
this_filepath <- paste0(this_folder, this_file)
df_plot <- read.csv(this_filepath, stringsAsFactors = F)
df_plot$X <- NULL
colnames(df_plot)[3:4] <- c("GDP / capita (USD), logged",
                            "Real expenditures per capita (USD), logged")
#--------------------------------------------------------------------------
mod <- lm(`Real expenditures per capita (USD), logged` ~ `GDP / capita (USD), logged`, df_plot)
#summary(mod)
m <- round(mod$coefficients[2], 2)
b <- round(mod$coefficients[1], 2)

# unique(df_plot$Country[which(df_plot$Region == "Asia\n(excluding W. Asia)")])
# unique(df_plot$Country[which(df_plot$Region == "N. Africa / W. Asia")])

df_plot$label_these <- NA
u <- df_plot$Country
df_plot$label_these[which(u %in% labelThese_vec)] <- u[which(u %in% labelThese_vec)]

n <- length(unique(df_plot$Region))
bag_of_colors <- randomcoloR::distinctColorPalette(k = 2 * n)
color_vec <- sample(bag_of_colors, n)

shape_vec <- c(21:24, 4)

gg <- ggplot(df_plot, aes(x = `GDP / capita (USD), logged`,
                          y = `Real expenditures per capita (USD), logged`,
                          group = Region, fill = Region, shape = Region,
                          label = label_these))
# gg <- gg + geom_point(aes(fill = Region, group = Region, shape = Region), pch = 21, alpha = 0.5, size = 6, color = "black")
gg <- gg + geom_point(alpha = 0.6, size = point_size, color = "black", stroke = 1)
gg <- gg + scale_fill_manual(values = color_vec)
gg <- gg + scale_shape_manual(values = shape_vec)
# gg <- gg + geom_abline(slope = m, intercept = b,
#                        linetype = "dashed", size = 1)
gg <- gg + geom_smooth(aes(group = NULL, fill = NULL, shape = NULL), method = lm, se = F)
gg <- gg + geom_text_repel(color = "black", size = label_size)
gg <- gg + labs(#title = "Engel's Law", 
  subtitle = paste0("Slope = ", m, ", y-intercept = ", b))#,
#caption = "Source: Author's creation based on World Bank International Comparison Program Data for 2017.
#GDP/capita is based on 2015 prices.")
gg <- gg + guides(fill = guide_legend(nrow = 2, byrow = T, override.aes = list(linetype = 0)),
                  color = guide_legend(override.aes = list(linetype = 0)))
gg <- gg + theme(legend.position = "bottom",
                 legend.spacing.x = unit(0.5, 'cm'),
                 legend.title = element_blank())
gg
#ggsave("Engels Law 2.png", width = 8.5, height = 6.5, units = "in")

```

Variation 1 disaggregated by region...again note how the fitted parameters vary from one region to another... In this case, Sub-Saharan Africa is more cleanly fit while the other regions are more poorly fit Figure \ref(fig:Engel1Disagg).

```{r Engel1Disagg, fig.show = "hold", fig.width = 7, fig.height=5, fig.align="center", fig.cap="Engel's Law, variation 1, by region, ICP data for 2017.", echo = FALSE}
region_vec <- unique(df_plot$Region)
facet_labels <- c()
for(i in 1:length(region_vec)){
  this_region <- region_vec[i]
  this_df_plot <- subset(df_plot, Region == this_region)
  mod <- lm(`Real expenditures per capita (USD), logged` ~ `GDP / capita (USD), logged`, this_df_plot)
  #summary(mod)
  m <- round(mod$coefficients[2], 2)
  b <- round(mod$coefficients[1], 2)
  #   print(this_region)
  # print(m)
  facet_labels[i] <- paste0("Slope = ", m, ", y-intercept = ", b)
}

names(facet_labels) <- region_vec

gg <- ggplot(df_plot, aes(x = `GDP / capita (USD), logged`,
                          y = `Real expenditures per capita (USD), logged`,
                          group = Region, fill = Region, shape = Region,
                          label = label_these))
gg <- gg + geom_point(alpha = 0.6, size = point_size, color = "black", stroke = 1)
gg <- gg + scale_fill_manual(values = color_vec)
gg <- gg + scale_shape_manual(values = shape_vec)
gg <- gg + geom_smooth(method = lm, se = F)
gg <- gg + facet_wrap(~Region, ncol = 3, scales = "free", labeller = labeller(Region = facet_labels))
gg <- gg + geom_text_repel(color = "black", size = label_size)
gg <- gg + theme(strip.background = element_blank(),
                 strip.text.x = element_text(hjust = -0.01),
                 #legend.direction = "vertical",
                 #legend.spacing.y = unit(1, 'cm'),
                 legend.title = element_blank())
gg <- gg + guides(fill = guide_legend(#nrow = 2,
  override.aes = list(linetype = 0)),
  color = guide_legend(override.aes = list(linetype = 0)))
gg <- shift_legend2(gg)
gg <- ggplotify::as.ggplot(gg)
#gg
#ggsave("Engels Law 2 Disaggregated.png", width = 8.5, height = 6.5, units = "in")
```

Tilman et al. [-@tilman2011global] examined a second variation of Engel's Law and used it to calculate future global food demand.^[Tilman et al. do not mention Engel's Law in their paper, believing this empirical relation to be newly discovered by them.] ...using FAO data...\ref(fig:Engel2). ...This is also useful since the FAO data extend back to the 1960s, whereas the ICP data extend back only to 2011, and only 2017 for Africa. Therefore it is possible to examine the stability of the empirical relation over a substantial period of time.

```{r Engel2, fig.show = "hold", fig.width = 5, fig.height=5, fig.align="center", fig.cap="Engel's Law, variation 2, FAO data for 2017.", echo = FALSE}
rm(df_plot)
this_file <- "Engels Law 3.csv"
this_filepath <- paste0(this_folder, this_file)
df_engel <- read.csv(this_filepath, stringsAsFactors = F)
df_engel$X <- NULL
colnames(df_engel)[3] <- "Food supply (kcal/capita/day), logged"
colnames(df_engel)[5] <- "GDP / capita (USD), logged"
#--------------------------------------------------------------------------
df_plot <- subset(df_engel, Year == 2017)
df_mod <- df_plot[, setdiff(colnames(df_plot), c("Country", "Year", "Region"))]
#keep_rows <- which(!is.na(df_mod$`Consumption Inequality (Gini Coefficient)`))
#df_mod <- df_mod[keep_rows, ]
keep_rows <- which(!is.na(df_mod$`Food supply (kcal/capita/day), logged`))
df_mod <- df_mod[keep_rows, ]

row.names(df_mod) <- df_plot$Country[keep_rows]
#df_mod$`Income Inequality (Gini Coefficient)` <- NULL
#df_mod$`Consumption Inequality (Gini Coefficient)` <- NULL
#df_mod$`Income Inequality (Gini Coefficient)` <- log(df_mod$`Income Inequality (Gini Coefficient)`)
#df_mod$`Consumption Inequality (Gini Coefficient)` <- log(df_mod$`Consumption Inequality (Gini Coefficient)`)

#df_mod$`Consumption Inequality (Gini Coefficient)` <- NULL

mod <- lm(`Food supply (kcal/capita/day), logged` ~ ., df_mod)
#summary(mod)
#m_gini <- round(mod$coefficients[3], 2)
m_gdp <- round(mod$coefficients[2], 2)
b <- round(mod$coefficients[1], 2)

# plot(mod$fitted.values, mod$residuals)
# plot(mod$fitted.values, df_mod$`Food supply (kcal/capita/day), logged`)

# y_lab <- paste(b, "+", m_gdp, "(GDP/capita (USD, 2015 prices), logged) +\n",
#       m_gini, "(Consumption Inequality (Gini Coefficient), logged)")

# df_mod$`6.93 + 0.1 (GDP/capita (USD, 2015 prices), logged) +\n -0.14 (Consumption Inequality (Gini Coefficient), logged)` <-
#   mod$fitted.values
# df_mod$Area <- row.names(df_mod)
# df_mod <- merge(df_mod, df_plot[, c("Area", "Region")], by = "Area")
# df_plot <- df_mod
#unique(df_plot$Area[grep("Korea", df_plot$Area, ignore.case = T)])
df_plot$label_these <- NA
u <- which(df_plot$Country %in% labelThese_vec)
df_plot$label_these[u] <- df_plot$Country[u]

# n <- length(unique(df_plot$Region))
# bag_of_colors <- randomcoloR::distinctColorPalette(k = 2 * n)
# color_vec <- sample(bag_of_colors, n)
gg <- ggplot(df_plot, aes(x = `GDP / capita (USD), logged`,
                          y = `Food supply (kcal/capita/day), logged`,
                          group = Region, fill = Region, shape = Region,
                          label = label_these))
# gg <- gg + geom_point(aes(fill = Region, group = Region, shape = Region), pch = 21, alpha = 0.5, size = 6, color = "black")
gg <- gg + geom_point(alpha = 0.6, size = point_size, color = "black", stroke = 1)
gg <- gg + scale_fill_manual(values = color_vec)
gg <- gg + scale_shape_manual(values = shape_vec)
# gg <- gg + geom_abline(slope = m, intercept = b,
#                        linetype = "dashed", size = 1)
gg <- gg + geom_smooth(aes(group = NULL, fill = NULL, shape = NULL), method = lm, se = F)
gg <- gg + geom_text_repel(color = "black", size = label_size)
gg <- gg + labs(#title = "Engel's Law", 
  subtitle = paste0("Slope = ", m_gdp, ", y-intercept = ", b))#,
#caption = "Source: Author's creation based on World Bank International Comparison Program Data for 2017.
#GDP/capita is based on 2015 prices.")
gg <- gg + guides(fill = guide_legend(nrow = 2, byrow = T, override.aes = list(linetype = 0)),
                  color = guide_legend(override.aes = list(linetype = 0)))
gg <- gg + theme(legend.position = "bottom",
                 legend.spacing.x = unit(0.5, 'cm'),
                 legend.title = element_blank())
gg
#ggsave("Engels Law 3.png", width = 8.5, height = 6.5, units = "in")


```

Perhaps the main benefit of this variation on Engel's Law is that the data go back a long ways, so you can investigate its stability over time... Figure \ref(fig:EngelAnim). The model Figure \ref(fig:EngelStability). An animation is also provided in the supplemental materials.

# Two laws, or two manifestations of the same law?

This is essentially a "production side" analogue of the time honored, empirical consumer side observation going back to Bernoulli (and perhaps as far back as Aristotle [@kauder1953genesis]) that _"in the absence of the unusual, the utility resulting from any small increase in wealth will be inversely proportionate to the quantity of goods previously possessed"_ [@bernoulli1954exposition].

Formalization of such statements depends upon how one interprets, mathematically, the notion of change expressed in words like "additional return" and "small increase". If these are interpreted as percentage changes, then the production side statement can be formalized as follows.

\begin{equation}
\frac{\partial \ln(R)}{\partial \ln(w_i)} \sim \frac{1}{w_i}
\label{eq:dimRetDefine}
\end{equation}

Where, to be clear, the term $\frac{\partial \ln(R)}{\partial \ln(w_i)}$ is the portfolio return elasticity with respect to investment in the $i^{th}$ portfolio item. In other words, this indicates the percentage increase in portfolio return given a $1$ percent increase in investment in the $i^{th}$ portfolio item. The subsequent portion of the equation, "$\sim 1 / w_i$", then means "inversely proportionate to the quantity of funds already invested".

Bernoulli [@...] law of diminishing returns to marginal increments... graph of utility curve... This implies that whether a good is inferior or superior depends on the quantity held.

As far as expenditures go, food stands to luxury goods as starchy foods stand to protein and vitamin rich foods.

# A new empirical law: expected price and price volatility

...Figure \ref{fig:price_series}. It seems that the higher the price, the more volatile it is... The phenomenon can be observed across a wide range of food and non-food agricultural commodities. It can also be seen to occur with both raw and processed versions of the commodities. The relation appears to persist across a wide range of food price epochs, from the oil driven spike of the 1970s through to the more recent 2007-2008 crisis. Plotting the mean price against its standard deviation offers striking confirmation... \ref{fig:risk_reward}

```{r priceSeries, fig.show = "hold", fig.width = 6, fig.height=4, fig.align="center", fig.cap="The higher the price, the more volatile. FAO data 1972-2017.", echo = FALSE}
this_file <- "FAO export data world.csv"
this_filepath <- paste0(this_folder, this_file)
df_export_world <- read.csv(this_filepath, stringsAsFactors = F)
df_export_world$X <- NULL
colnames(df_export_world)[4:6] <- c("Export Quantity", "Export Value",
                                    "World Export Price (USD / metric ton)")
#=============================================================================
plot_these <- c("Maize", "Apples", "Cocoa, beans", "Cocoa, butter")
df_plot <- subset(df_export_world, Item %in% plot_these &
                    Year >= 1972)
yr_1 <- unique(df_plot$Year)[1]
yr_n <- max(unique(df_plot$Year))
gg <- ggplot(df_plot, aes(x = Year, y = `World Export Price (USD / metric ton)`, group = Item, color = Item)) + geom_line(lwd = 1)
gg <- gg + scale_x_continuous(breaks = seq(yr_1, yr_n, length.out = 6))
#gg <- gg + scale_y_log10()
gg <- gg + theme(axis.title.x = element_blank(),
                 legend.title = element_blank(),
                 legend.position = "top")
gg

```

...Figure \@ref(fig:riskReward)

```{r riskReward, fig.show = "hold", fig.width = 6, fig.height=6, fig.align="center", fig.cap="Logged expected price plotted against logged standard deviation. 'Export price' here equals FAO's Export Value series divided by its Export Quantity series, downloaded from FAOSTAT.", echo = FALSE}

df_riskReward <- df_export_world %>% group_by(Item) %>%
  summarise(`Mean World Export Price (USD / metric ton)` = mean(`World Export Price (USD / metric ton)`, na.rm = T),
            `Stand. Dev. World Export Price (USD / metric ton)` = sd(`World Export Price (USD / metric ton)`, na.rm = T),
            `Mean Export Qty.` = mean(`Export Quantity`, na.rm = T), 
            `Mean Export Value ('000 USD)` = mean(`Export Value`, na.rm = T))

df_riskReward$`Mean World Export Price (USD / metric ton), logged` <-
  log(df_riskReward$`Mean World Export Price (USD / metric ton)`)
df_riskReward$`Stand. Dev. World Export Price (USD / metric ton), logged` <-
  log(df_riskReward$`Stand. Dev. World Export Price (USD / metric ton)`)
#--
df_riskReward <- df_riskReward[which(!is.na(df_riskReward$`Stand. Dev. World Export Price (USD / metric ton), logged`)), ]
#nrow(df_riskReward)
#----------------------------------------------------------------------------
df_mod <- df_riskReward[, c("Mean World Export Price (USD / metric ton), logged", "Stand. Dev. World Export Price (USD / metric ton), logged")]
#nrow(df_mod)
row.names(df_mod) <- df_riskReward$Item
mod <- lm(`Mean World Export Price (USD / metric ton), logged` ~., df_mod)
#summary(mod)
b <- round(mod$coefficients[1], 2)
m <- round(mod$coefficients[2], 2)
#---
# gg <- ggplot(data.frame(Item = row.names(df_mod), yhat = mod$fitted.values,
#                         resid = mod$residuals),
#              aes(x = yhat, y = resid, label = Item))
# gg <- gg + geom_point() + geom_text()
# gg <- gg + geom_hline(yintercept = c(-1, 1), color = "red")
# gg
#---
ind_outlier <- which(abs(as.numeric(mod$residuals)) >= 0.75)
excluded_items <- row.names(df_mod)[ind_outlier]
pct_removed <- 100 * length(ind_outlier) / nrow(df_riskReward)
row_names <- row.names(df_mod)[-ind_outlier]
df_mod <- df_mod[-ind_outlier, ]
row.names(df_mod) <- row_names
mod <- lm(`Mean World Export Price (USD / metric ton), logged` ~., df_mod)
#summary(mod)
b <- round(mod$coefficients[1], 2)
m <- round(mod$coefficients[2], 2)
#plot(mod$fitted.values, mod$residuals)
#----------------------------------------------------------------------------
#df_riskReward$Item[ind_outlier]
df_plot <- df_riskReward[-ind_outlier, c("Item", "Mean World Export Price (USD / metric ton), logged", "Stand. Dev. World Export Price (USD / metric ton), logged",
                             "Mean Export Value ('000 USD)")]

df_plot$`Mean Export Value\n(million USD)` <- df_plot$`Mean Export Value ('000 USD)` / 1000

labelItems <- c("Maize", "Cassava dried", "Flour, wheat", "Starch, cassava",
              "Coffee, green", "Coffee, roasted", "Wheat", "Bananas",
              "Plantains", "Potatoes", "Cassava Equivalent", "Cocoa, beans",
              "Wool, greasy", "Beans, dry", "Cotton lint", "Silk", "Apples",
              "Soybeans", "Oil, soybean", "Rubber, natural", "Rice",
              "Cocoa, butter")

df_plot$label_these <- NA
u <- df_plot$Item
df_plot$label_these[which(u %in% labelItems)] <- u[which(u %in% labelItems)]

label_size <- 4
gg <- ggplot(df_plot, aes(x = `Stand. Dev. World Export Price (USD / metric ton), logged`,
                          y = `Mean World Export Price (USD / metric ton), logged`))
gg <- gg + geom_smooth(method = lm, se = F)
gg <- gg + geom_point(aes(size = `Mean Export Value\n(million USD)`),
                      fill = "khaki1", pch=21, color="Black")
gg <- gg + geom_text_repel(aes(label = label_these), size = label_size)
gg <- gg + labs(title = "The risk-return frontier", subtitle = paste0("Slope = ", m, "  Y intercept = ", b))
gg <- gg + theme(legend.position = "bottom")
gg <- gg + labs(title = "The ag commodity risk-return frontier",
                caption = "Source: Author's creation using FAOSTAT export value and quantity series 1961-2017.")
gg
#ggsave("risk reward frontier.pdf")
```

...Figure \ref{fig:riskRewardByRegion}

```{r riskRewardByRegion, fig.show = "hold", fig.width = 7, fig.height=6, fig.align="center", fig.cap="Logged expected price plotted against logged standard deviation, by region.", echo = FALSE}
this_file <- "FAO export data regions.csv"
this_filepath <- paste0(this_folder, this_file)
df_export_regions <- read.csv(this_filepath, stringsAsFactors = F)
df_export_regions$X <- NULL
colnames(df_export_regions)[4:6] <- c("Export Quantity", "Export Value",
                                    "Region Export Price (USD / metric ton)")
#============================================================================
#colnames(df_export_regions)
df_riskReward_regions <- df_export_regions %>% group_by(Region, Item) %>%
  summarise(`Mean Region Export Price (USD / metric ton)` = mean(`Region Export Price (USD / metric ton)`, na.rm = T),
            `Stand. Dev. Region Export Price (USD / metric ton)` = sd(`Region Export Price (USD / metric ton)`, na.rm = T),
            `Mean Export Qty.` = mean(`Export Quantity`, na.rm = T), 
            `Mean Export Value ('000 USD)` = mean(`Export Value`, na.rm = T))

df_riskReward_regions$`Mean Region Export Price (USD / metric ton), logged` <-
  log(df_riskReward_regions$`Mean Region Export Price (USD / metric ton)`)
df_riskReward_regions$`Stand. Dev. Region Export Price (USD / metric ton), logged` <-
  log(df_riskReward_regions$`Stand. Dev. Region Export Price (USD / metric ton)`)
df_riskReward_regions$`Mean Export Value\n(million USD)` <-
  df_riskReward_regions$`Mean Export Value ('000 USD)` / 1000
#--
df_riskReward_regions <- df_riskReward_regions[which(!is.na(df_riskReward_regions$`Stand. Dev. Region Export Price (USD / metric ton), logged`)), ]
ind_rm <- which(is.infinite(df_riskReward_regions$`Stand. Dev. Region Export Price (USD / metric ton), logged`))
#df_riskReward_regions$Item[ind_rm] #look
df_riskReward_regions <- df_riskReward_regions[-ind_rm, ]
#nrow(df_riskReward_regions)
#----------------------------------------------------------------------------
df_plot <- df_riskReward_regions
gg <- ggplot(df_plot, aes(x = `Stand. Dev. Region Export Price (USD / metric ton), logged`,
                          y = `Mean Region Export Price (USD / metric ton), logged`))
#gg <- gg + geom_smooth(method = lm, se = F)
gg <- gg + geom_point(aes(size = `Mean Export Value\n(million USD)`),
                      fill = "khaki1", pch=21, color="Black")
#gg <- gg + geom_text_repel(aes(label = label_these), size = label_size)
gg <- gg + facet_wrap(~Region)
gg <- gg + labs(title = "The ag commodity risk-return frontier, by region",
                caption = "Source: Author's creation using FAOSTAT export value and quantity series 1961-2017.")
gg <- shift_legend2(gg)
gg <- ggplotify::as.ggplot(gg)
```



































































```{r, fig.show = "hold", fig.width = 5, fig.height=5, fig.align='left', fig.cap="\\label{fig:Engel0}Engel's Law. Source: Author's creation based on ICP data for 2017. GDP/capita is retrieved from the FAO database and is based on 2015 prices.", echo = FALSE}
#===========================================================================
FAOdat_createRegionGroups <- function(df_raw, exclude_these, keep_FAOregions = F,
                                      df_is_trade_matrix = F,
                                      consolidate_LAC = T,
                                      consolidate_WEur = T,
                                      consolidate_AusNZPacIslands = T,
                                      consolidate_SSA = T,
                                      output_countryVecs = F)
{
  #=================================
  #--Africa
  countries_NAfrica <- as.character(unique(read.csv("Country list - Northern Africa.csv")[,"Area"]))
  countries_MAfrica <- as.character(unique(read.csv("Country list - Middle Africa.csv")[,"Area"]))
  countries_WAfrica <- as.character(unique(read.csv("Country list - Western Africa.csv")[,"Area"]))
  countries_EAfrica <- as.character(unique(read.csv("Country list - Eastern Africa.csv")[,"Area"]))
  countries_SAfrica <- as.character(unique(read.csv("Country list - Southern Africa.csv")[,"Area"]))
  #--Americas
  countries_SAmer <- as.character(unique(read.csv("Country list - South America.csv")[,"Area"]))
  countries_CAmer <- as.character(unique(read.csv("Country list - Central America.csv")[,"Area"]))
  countries_Carib <- as.character(unique(read.csv("Country list - Caribbean.csv")[,"Area"]))
  countries_NAmer <- as.character(unique(read.csv("Country list - Northern America.csv")[,"Country"]))
  #--Asia
  countries_EAsia <- as.character(unique(read.csv("Country list - Eastern Asia.csv")[,"Area"]))
  #countries_EAsia <- countries_EAsia[!(countries_EAsia %in% c("China, Hong Kong SAR", "China, Macao SAR"))]
  countries_SEAsia <- as.character(unique(read.csv("Country list - South-Eastern Asia.csv")[,"Area"]))
  countries_SAsia <- as.character(unique(read.csv("Country list - Southern Asia.csv")[,"Area"]))
  countries_WAsia <- as.character(unique(read.csv("Country list - Western Asia.csv")[,"Area"]))
  countries_CAsia <- as.character(unique(read.csv("Country list - Central Asia.csv")[,"Area"]))
  #--Europe
  countries_NEurope <- as.character(unique(read.csv("Country list - Northern Europe.csv")[,"Area"]))
  countries_WEurope <- as.character(unique(read.csv("Country list - Western Europe.csv")[,"Area"]))
  countries_EEurope <- as.character(unique(read.csv("Country list - Eastern Europe.csv")[,"Area"]))
  countries_SEurope <- as.character(unique(read.csv("Country list - Southern Europe.csv")[,"Area"]))
  #--Oceania
  countries_Oceania <- as.character(unique(read.csv("Country list - Oceania.csv")[,"Area"]))
  countries_AusNZea <- c("Australia", "New Zealand")
  countries_PacifIs <- setdiff(countries_Oceania, countries_AusNZea)
  #=================================
  #--Create region groupings
  if(df_is_trade_matrix){
    u <- df_raw$Partner.Countries
  }else{
    u <- df_raw$Area
  }
  df_raw$Region <- NA
  df_raw$Region[which(u %in% countries_NAmer)] <- "North America"
  df_raw$Region[which(u %in% countries_SAmer)] <- "South America"
  df_raw$Region[which(u %in% countries_CAmer)] <- "Central America"
  df_raw$Region[which(u %in% countries_Carib)] <- "Caribbean"
  df_raw$Region[which(u %in% countries_NAfrica)] <- "Northern Africa"
  df_raw$Region[which(u %in% countries_SAfrica)] <- "Southern Africa"
  df_raw$Region[which(u %in% countries_WAfrica)] <- "Western Africa"
  df_raw$Region[which(u %in% countries_EAfrica)] <- "Eastern Africa"
  df_raw$Region[which(u %in% countries_MAfrica)] <- "Middle Africa"
  df_raw$Region[which(u %in% countries_CAsia)] <- "Central Asia"
  df_raw$Region[which(u %in% countries_WAsia)] <- "Western Asia"
  df_raw$Region[which(u %in% countries_SAsia)] <- "Southern Asia"
  df_raw$Region[which(u %in% countries_EAsia)] <- "Eastern Asia"
  df_raw$Region[which(u %in% countries_SEAsia)] <- "South-Eastern Asia"
  df_raw$Region[which(u %in% countries_NEurope)] <- "Northern Europe"
  df_raw$Region[which(u %in% countries_SEurope)] <- "Southern Europe"
  df_raw$Region[which(u %in% countries_WEurope)] <- "Western Europe"
  df_raw$Region[which(u %in% countries_EEurope)] <- "Eastern Europe"
  df_raw$Region[which(u %in% countries_PacifIs)] <- "Pacific Islands"
  df_raw$Region[which(u %in% countries_AusNZea)] <- "Australia & New Zealand"
  rm(u)
  #--------------------
  #--See what countries escaped designation
  #unique(df_raw$Area[which(is.na(df_raw$Region))])
  #--Assign these to their proper regions
  #(Leave out "China" as it is already covered under "China, mainlaind", "Hong Kong", etc.)
  if(df_is_trade_matrix){
    u <- df_raw$Partner.Countries
  }else{
    u <- df_raw$Area
  }
  df_raw$Region[which(u %in% c("?land Islands", "Isle of Man", "Greenland"))] <- "Northern Europe"
  df_raw$Region[which(u %in% c("Anguilla", "Bermuda", "Cayman Islands", "Cura?ao"))] <- "Caribbean"
  df_raw$Region[which(u %in% c("C?te d'Ivoire"))] <- "Western Africa"
  df_raw$Region[which(u %in% c("Palau"))] <- "Pacific Islands"
  df_raw$Region[which(u %in% c("Maldives", "R?union"))] <- "Southern Asia"
  df_raw$Region[which(u %in% c("French Guiana"))] <- "South America"
  rm(u)
  #===========================
  #unique(df_raw$Area[which(is.na(df_raw$Region) == T)])
  if(!(df_is_trade_matrix)){
    ind_NA <- which(is.na(df_raw$Region) == T)
    ind_noNA <- which(is.na(df_raw$Region) == F)
    if(keep_FAOregions == F){
      df_raw <- df_raw[ind_noNA,]
    }else{
      df_raw$Region[ind_NA] <- df_raw$Area[ind_NA] 
    }
  }
  #===========================
  #--
  if(df_is_trade_matrix){
    df_raw <- subset(df_raw, !(Partner.Countries %in% exclude_these))
  }else{
    df_raw <- subset(df_raw, !(Area %in% exclude_these))
  }
  #--
  if(consolidate_LAC){
    LAC <- c("Central America", "Caribbean", "South America")
    u <- df_raw$Region
    df_raw$Region[which(u %in% LAC)] <- "LAC"
    rm(u)
  }
  if(consolidate_WEur){
    Europe_WNS <- c("Southern Europe", "Western Europe", "Northern Europe")
    u <- df_raw$Region
    df_raw$Region[which(u %in% Europe_WNS)] <- "W/N/S Europe"
    rm(u)
  }
  if(consolidate_AusNZPacIslands){
    AusNZPacIslands <- c("Australia & New Zealand", "Pacific Islands")
    u <- df_raw$Region
    df_raw$Region[which(u %in% AusNZPacIslands)] <- "Aus/NZ/Oceania"
    rm(u)
  }
  if(consolidate_SSA){
    SSA <- c("Eastern Africa", "Southern Africa", "Western Africa", "Middle Africa")
    u <- df_raw$Region
    df_raw$Region[which(u %in% SSA)] <- "Sub-Saharan Africa"
    rm(u)
  }
  
  # Europe_E <- "Eastern Europe"
  # ESE_Asia <- c("South-Eastern Asia", "Eastern Asia")
  #--
  # u <- df_raw$Region
  # df_raw$Region[which(u %in% LAC)] <- "LAC"
  # df_raw$Region[which(u %in% Europe_E)] <- "E. Europe"
  # df_raw$Region[which(u %in% ESE_Asia)] <- "E. & S.E. Asia"
  # rm(u)
  #--
  if(output_countryVecs){
    out_list <- list()
    out_list[["Northern Africa"]] <- countries_NAfrica
    out_list[["Middle Africa"]] <- countries_MAfrica
    out_list[["Western Africa"]] <- countries_WAfrica
    out_list[["Eastern Africa"]] <- countries_EAfrica
    out_list[["Southern Africa"]] <- countries_SAfrica
    out_list[["df_raw"]] <- df_raw
    return(out_list)
  }else{
    return(df_raw)
  }
  
}

#===========================================================================
df_icp_raw <- read.csv("WB_ICP_All.csv", stringsAsFactors = F)
#colnames(df_icp_raw)
rm_cols <- c("Country.Code", "Classification.Code", "Series.Code")
rm_cols <- which(colnames(df_icp_raw) %in% rm_cols)
df_icp_raw <- df_icp_raw[, -c(rm_cols)]
#colnames(df)
u <- colnames(df_icp_raw)
colnames(df_icp_raw) <- gsub("^.*YR", "", u)
u <- colnames(df_icp_raw)
colnames(df_icp_raw) <- gsub("\\.", "", u)
# unique(df_icp_raw$ClassificationName)
# unique(df_icp_raw$SeriesName)
colnames(df_icp_raw)[1:3] <- c("Country", "Element", "Item")
gather_these <- colnames(df_icp_raw)[4:ncol(df_icp_raw)]
df_icp_raw <- df_icp_raw %>% gather_("Year", "Value", gather_these)
df_icp_raw$Year <- as.integer(df_icp_raw$Year)
df_icp_raw$Value <- as.numeric(df_icp_raw$Value)
u <- df_icp_raw$Item
df_icp_raw$Item <- gsub("^.*:", "", u)
#unique(df_icp_raw$Country)
df_icp_raw <- subset(df_icp_raw, !(Country %in% c("", "Data from database: ICP 2017",
                                                  "Last Updated: 06/19/2020")))
df_icp_raw$Country[which(df_icp_raw$Country == "Korea, Rep.")] <- "South Korea"
df_icp_raw$Country[grep("Congo, Dem. Rep.", df_icp_raw$Country)] <- "DRC"
df_icp_raw$Country[grep("Congo", df_icp_raw$Country)] <- "Congo"
df_icp_raw$Country[grep("Cte d'Ivoire", df_icp_raw$Country)] <- "C?te d'Ivoire"
df_icp_raw$Country[grep("Egypt", df_icp_raw$Country)] <- "Egypt"
df_icp_raw$Country[grep("Curaao", df_icp_raw$Country)] <- "Cura?ao"
df_icp_raw$Country[grep("So Tom and Principe", df_icp_raw$Country)] <- "Sao Tome and Principe"
df_icp_raw$Country <- gsub("St\\.", "Saint", df_icp_raw$Country)
df_icp_raw$Country[grep("Iran", df_icp_raw$Country)] <- "Iran"
df_icp_raw$Country[grep("Gambia", df_icp_raw$Country)] <- "Gambia"
df_icp_raw$Country[grep("Bahamas", df_icp_raw$Country)] <- "Bahamas"
df_icp_raw$Country[grep("Venezuela", df_icp_raw$Country)] <- "Venezuela"
df_icp_raw$Country[grep("Cape Verde", df_icp_raw$Country)] <- "Cabo Verde"
df_icp_raw$Country[grep("Virgin Islands, British", df_icp_raw$Country)] <- "British Virgin Islands"
#unique(df_icp_raw$Country[grep("Guatemala", df_icp_raw$Country)])
#unique(df_gdp_raw$Area[grep("China", df_gdp_raw$Area)])
#unique(df_icp_raw$Item)
#unique(df_icp_raw$Element)
df_icp <- subset(df_icp_raw, Item == "FOOD AND NON-ALCOHOLIC BEVERAGES" &
                   Element == "Expenditure shares (GDP = 100)" &
                   Year == 2017)
df_icp <- df_icp[which(!is.na(df_icp$Value)), ]
df_icp <- df_icp[, c("Country", "Value")]
df_icp$Value <- log(df_icp$Value)
colnames(df_icp)[ncol(df_icp)] <- "Food expenditure share of GDP, logged"
#---------------------------------------------------------------------------
df_gdp_raw <- read.csv("Macro-Statistics_Key_Indicators_E_All_Data.csv", stringsAsFactors = F)
rm_cols <- c("Area.Code", "Item.Code", "Element.Code")
rm_cols <- which(colnames(df_gdp_raw) %in% rm_cols)
df_gdp_raw <- df_gdp_raw[, -rm_cols]
#unique(df_gdp_raw$Item)
df_gdp_raw$Item <- as.character(df_gdp_raw$Item)
df_gdp_raw$Area <- as.character(df_gdp_raw$Area)
u <- colnames(df_gdp_raw)
df_gdp_raw <- df_gdp_raw[, -grep("F", u)]
u <- colnames(df_gdp_raw)
df_gdp_raw <- df_gdp_raw[, -grep("N", u)]
v <- colnames(df_gdp_raw)[5:ncol(df_gdp_raw)]
colnames(df_gdp_raw)[5:ncol(df_gdp_raw)] <- gsub("Y", "", v)
#---------------------------------------------------------------------------
# Check available years
#colnames(df_gdp_raw)
available_yrs <- as.character(c(1970:2018))
df_gdp_raw <- gather_(df_gdp_raw, "Year", "Value", gather_cols = available_yrs)
df_gdp_raw <- df_gdp_raw
#df_gdp_raw$Unit <- NULL
#---------------------------------------------------------------------------
out_list <- FAOdat_createRegionGroups(df_gdp_raw, exclude_these = NULL,
                                      keep_FAOregions = F,
                                      df_is_trade_matrix = F,
                                      consolidate_LAC = T,
                                      consolidate_WEur = T,
                                      consolidate_AusNZPacIslands = T,
                                      consolidate_SSA = F,
                                      output_countryVecs = T)
df_gdp_raw <- out_list[["df_raw"]]
#unique(df_gdp_raw$Area[which(is.na(df_gdp_raw$Region))])
#unique(df_gdp_raw$Area[grep("Guatemala", df_gdp_raw$Area)])
#unique(df_gdp_raw$Region[grep("d'Ivoire", df_gdp_raw$Area)])
#---------------------------------------------------------------------------
df_gdp_raw$Area[grep("Tanzania", df_gdp_raw$Area)] <- "Tanzania"
#df_gdp_raw$Area[grep("Eswatini", df_gdp_raw$Area)] <- "Swaziland"
df_gdp_raw$Area[grep("Cte d'Ivoire", df_gdp_raw$Area)] <- "C?te d'Ivoire"
df_gdp_raw$Area[grep("Viet Nam", df_gdp_raw$Area)] <- "Vietnam"
df_gdp_raw$Area[grep("Republic of the Congo", df_gdp_raw$Area)] <- "DRC"
df_gdp_raw$Area[grep("Congo", df_gdp_raw$Area)] <- "Congo"
df_gdp_raw$Area[grep("United States of America", df_gdp_raw$Area)] <- "United States"
df_gdp_raw$Area[which(df_gdp_raw$Area == "Republic of Korea")] <- "South Korea"
df_gdp_raw$Area[grep("Lao People's Democratic", df_gdp_raw$Area)] <- "Lao PDR"
df_gdp_raw$Area[grep("Venezuela", df_gdp_raw$Area)] <- "Venezuela"
df_gdp_raw$Area[grep("Bolivia", df_gdp_raw$Area)] <- "Bolivia"
df_gdp_raw$Area[grep("Sint Maarten", df_gdp_raw$Area)] <- "Sint Maarten"
df_gdp_raw$Area[grep("Iran", df_gdp_raw$Area)] <- "Iran"
df_gdp_raw$Area[grep("Czechia", df_gdp_raw$Area)] <- "Czech Republic"
df_gdp_raw$Area[grep("Moldova", df_gdp_raw$Area)] <- "Moldova"
df_gdp_raw$Area[grep("China, Hong Kong SAR", df_gdp_raw$Area)] <- "Hong Kong SAR, China"
#unique(df_gdp_raw$Area[grep("Cape", df_gdp_raw$Area)])
#df_gdp$Country[which(df_gdp$Country == "China, mainland")] <- "China"
#---------------------------------------------------------------------------
#unique(df_gdp_raw$Item)
item_vec <- c("Gross Domestic Product per capita")
this_element <- "Value US$, 2015 prices"
df_gdp <- subset(df_gdp_raw, Item == "Gross Domestic Product per capita" &
                   Element == "Value US$, 2015 prices" &
                   Year == 2017)
df_gdp <- df_gdp[, c("Area", "Region", "Value")]
colnames(df_gdp)[1] <- "Country"
df_gdp$Value <- log(df_gdp$Value)
colnames(df_gdp)[ncol(df_gdp)] <- "GDP/capita (USD, 2015 prices), logged"
#--------------------------------------------------------------------------
#df_icp$Country[which(df_icp$Country == "China")] <- "China, mainland"
#unique(df_icp$Country[grep("Timor-Leste", df_icp$Country)])
dont_match1 <- setdiff(unique(df_icp$Country), unique(df_gdp$Country))
dont_match2 <- setdiff(unique(df_gdp$Country), unique(df_icp$Country))
dont_match <- c(dont_match1, dont_match2)
df_engel1 <- merge(df_gdp, df_icp, by = c("Country"))
#==========================================================================
df_engel1$Region[grep("Western Asia", df_engel1$Region)] <- "N. Africa / W. Asia"
df_engel1$Region[grep("Northern Africa", df_engel1$Region)] <- "N. Africa / W. Asia"
df_engel1$Region[which(df_engel1$Region %in% c("Western Africa",
                                               "Eastern Africa",
                                               "Middle Africa",
                                               "Southern Africa"))] <- "Sub-Saharan Africa"
df_engel1$Region[grep("Europe", df_engel1$Region)] <- "Europe"
df_engel1$Region[which(df_engel1$Region %in% c("Europe", "North America"))] <- "Europe / N. Amer.\nAustralia / NZ"
df_engel1$Region[which(df_engel1$Area %in% c("Australia", "New Zealand"))] <- "Europe / N. Amer.\nAustralia / NZ"
df_engel1$Region[which(df_engel1$Region %in% c("Southern Asia", "South-Eastern Asia", "Central Asia", "Eastern Asia"))] <- "Asia\n(excluding W. Asia)"
df_engel1$Region[which(df_engel1$Region == "Aus/NZ/Oceania")] <- "Pacific Islands"
#unique(df_engel1$Region)
#-------------------------------------------------------------------------
df_plot <- df_engel1
colnames(df_plot)[3] <- c("GDP / capita (USD), logged")
df_plot <- subset(df_plot, Region != "Pacific Islands" &
                    !(Country %in% c("Saint Lucia", "Seychelles")))
#write.csv(df_plot, "Engels Law 1.csv", col.names = colnames(df_plot))
#--------------------------------------------------------------------------
mod <- lm(`Food expenditure share of GDP, logged` ~ `GDP / capita (USD), logged`, df_plot)
#summary(mod)
m <- round(mod$coefficients[2], 2)
b <- round(mod$coefficients[1], 2)

# unique(df_plot$Country[which(df_plot$Region == "Asia\n(excluding W. Asia)")])
# unique(df_plot$Country[which(df_plot$Region == "N. Africa / W. Asia")])
# label_Africa <- c("Tanzania", "Uganda", "Kenya", "Ethiopia", "South Africa",
#                   "DRC", "Angola", "Ghana","Rwanda", "Mozambique")
# label_NAfricaWAsia <- c("Egypt", "Turkey", "Sudan", "Saudi Arabia", "Jordan",
#                         "Israel", "United Arab Emirates", "Kuwait")
# label_Asia <- c("India", "Iran", "China, mainland", "Vietnam", "Japan",
#                 "South Korea", "Afghanistan", "Indonesia", "Kazakhstan",
#                 "Singapore", "Tajikistan")
# label_NAmEurAusNZ <- c("Austria", "United Kingdom", "France", "Australia",
#                        "United States", "Spain", "Italy")
# label_LAC <- c("Nicaragua", "Haiti", "Argentina", "Colombia", "Mexico",
#                "Dominican Republic", "Brazil", "Uruguay")
# labelThese_vec <- c(label_Africa, label_NAfricaWAsia, label_Asia,
#                     label_NAmEurAusNZ, label_LAC)
label_Africa <- c("Tanzania", "Uganda", "Ethiopia", "South Africa",
                  "Ghana")
label_NAfricaWAsia <- c("Turkey", "Saudi Arabia")
label_Asia <- c("India", "Iran", "China, mainland", "Vietnam", "Japan",
                "South Korea")
label_NAmEurAusNZ <- c("Italy", "France", "Australia")
label_LAC <- c("Nicaragua", "Colombia", "Brazil")
labelThese_vec <- c(label_Africa, label_NAfricaWAsia, label_Asia,
                    label_NAmEurAusNZ, label_LAC)

df_plot$label_these <- NA
u <- df_plot$Country
df_plot$label_these[which(u %in% labelThese_vec)] <- u[which(u %in% labelThese_vec)]

n <- length(unique(df_plot$Region))
bag_of_colors <- randomcoloR::distinctColorPalette(k = 2 * n)
color_vec <- sample(bag_of_colors, n)

shape_vec <- c(21:24, 4)
point_size <- 2
label_size <- 4

gg <- ggplot(df_plot, aes(x = `GDP / capita (USD), logged`,
                          y = `Food expenditure share of GDP, logged`,
                          group = Region, fill = Region, shape = Region,
                          label = label_these))
# gg <- gg + geom_point(aes(fill = Region, group = Region, shape = Region), pch = 21, alpha = 0.5, size = 6, color = "black")
gg <- gg + geom_point(alpha = 0.6, size = point_size, color = "black", stroke = 1)
gg <- gg + scale_fill_manual(values = color_vec)
gg <- gg + scale_shape_manual(values = shape_vec)
# gg <- gg + geom_abline(slope = m, intercept = b,
#                        linetype = "dashed", size = 1)
gg <- gg + geom_smooth(aes(group = NULL, fill = NULL, shape = NULL), method = lm, se = F)
gg <- gg + geom_text_repel(color = "black", size = label_size)
gg <- gg + labs(#title = "Engel's Law", 
  subtitle = paste0("Slope = ", m, ", y-intercept = ", b))#,
#caption = "Source: Author's creation based on World Bank International Comparison Program Data for 2017.
#GDP/capita is based on 2015 prices.")
gg <- gg + guides(fill = guide_legend(nrow = 2, byrow = T, override.aes = list(linetype = 0)),
                  color = guide_legend(override.aes = list(linetype = 0)))
gg <- gg + theme(legend.position = "bottom",
                 legend.spacing.x = unit(0.5, 'cm'),
                 legend.title = element_blank())
gg
#ggsave("Engels Law 1.png", width = 8.5, height = 6.5, units = "in")
```

Better fit when examined by regions except for Sub-Saharan Africa... Note how parameter values vary somewhat by region in Figure \ref{fig:Engel0_disagg}. Might also try disaggregating by quantile.

```{r, fig.show = "hold", fig.width = 5, fig.height=6, fig.align='left', fig.cap="\\label{fig:Engel0_disagg}Engel's Law by region, ICP data for 2017.", echo = FALSE}
region_vec <- unique(df_plot$Region)
facet_labels <- c()
for(i in 1:length(region_vec)){
  this_region <- region_vec[i]
  this_df_plot <- subset(df_plot, Region == this_region)
  mod <- lm(`Food expenditure share of GDP, logged` ~ `GDP / capita (USD), logged`, this_df_plot)
  #summary(mod)
  m <- round(mod$coefficients[2], 2)
  b <- round(mod$coefficients[1], 2)
  #   print(this_region)
  # print(m)
  facet_labels[i] <- paste0("Slope = ", m, ", y-intercept = ", b)
}

names(facet_labels) <- region_vec

gg <- ggplot(df_plot, aes(x = `GDP / capita (USD), logged`,
                          y = `Food expenditure share of GDP, logged`,
                          group = Region, fill = Region, shape = Region,
                          label = label_these))
gg <- gg + geom_point(alpha = 0.6, size = point_size, color = "black", stroke = 1)
gg <- gg + scale_fill_manual(values = color_vec)
gg <- gg + scale_shape_manual(values = shape_vec)
gg <- gg + geom_smooth(method = lm, se = F)
gg <- gg + facet_wrap(~Region, ncol = 2, scales = "free", labeller = labeller(Region = facet_labels))
gg <- gg + geom_text_repel(color = "black", size = label_size)
gg <- gg + theme(strip.background = element_blank(),
                 strip.text.x = element_text(hjust = -0.01),
                 #legend.direction = "vertical",
                 #legend.spacing.y = unit(1, 'cm'),
                 legend.title = element_blank())
gg <- gg + guides(fill = guide_legend(#nrow = 2,
  override.aes = list(linetype = 0)),
  color = guide_legend(override.aes = list(linetype = 0)))
gg <- shift_legend2(gg)
gg <- ggplotify::as.ggplot(gg)
#gg
#ggsave("Engels Law 1 Disaggregated.png", width = 8.5, height = 6.5, units = "in")
```


"Engel's law establishes that as income increases, households' demand for
food increases less than proportionally" [@cirera2010income]... Figure \ref{fig:Engel1}.


```{r, fig.show = "hold", fig.width = 5, fig.height=5, fig.align='left', fig.cap="\\label{fig:Engel1}Engel's Law, variation 1, ICP data for 2017.", echo = FALSE}

#unique(df_icp_raw$Element)
df_icp <- subset(df_icp_raw, Item == "FOOD AND NON-ALCOHOLIC BEVERAGES" &
                   Element == "Real expenditures per capita (U.S. dollars)" &
                   Year == 2017)
df_icp <- df_icp[which(!is.na(df_icp$Value)), ]
df_icp <- df_icp[, c("Country", "Value")]
df_icp$Value <- log(df_icp$Value)
colnames(df_icp)[ncol(df_icp)] <- "Real expenditures per capita (USD), logged"
#--------------------------------------------------------------------------
#df_icp$Country[which(df_icp$Country == "China")] <- "China, mainland"
#unique(df_icp$Country[grep("Timor-Leste", df_icp$Country)])
dont_match1 <- setdiff(unique(df_icp$Country), unique(df_gdp$Country))
dont_match2 <- setdiff(unique(df_gdp$Country), unique(df_icp$Country))
dont_match <- c(dont_match1, dont_match2)
df_engel <- merge(df_gdp, df_icp, by = c("Country"))
#==========================================================================
df_engel$Region[grep("Western Asia", df_engel$Region)] <- "N. Africa / W. Asia"
df_engel$Region[grep("Northern Africa", df_engel$Region)] <- "N. Africa / W. Asia"
df_engel$Region[which(df_engel$Region %in% c("Western Africa",
                                             "Eastern Africa",
                                             "Middle Africa",
                                             "Southern Africa"))] <- "Sub-Saharan Africa"
df_engel$Region[grep("Europe", df_engel$Region)] <- "Europe"
df_engel$Region[which(df_engel$Region %in% c("Europe", "North America"))] <- "Europe / N. Amer.\nAustralia / NZ"
df_engel$Region[which(df_engel$Area %in% c("Australia", "New Zealand"))] <- "Europe / N. Amer.\nAustralia / NZ"
df_engel$Region[which(df_engel$Region %in% c("Southern Asia", "South-Eastern Asia", "Central Asia", "Eastern Asia"))] <- "Asia\n(excluding W. Asia)"
df_engel$Region[which(df_engel$Region == "Aus/NZ/Oceania")] <- "Pacific Islands"
#unique(df_engel$Region)
#-------------------------------------------------------------------------
df_plot <- df_engel
colnames(df_plot)[3] <- c("GDP / capita (USD), logged")
df_plot <- subset(df_plot, Region != "Pacific Islands" &
                    !(Country %in% c("Saint Lucia", "Seychelles")))
#write.csv(df_plot, "Engels Law 2.csv")
#--------------------------------------------------------------------------
mod <- lm(`Real expenditures per capita (USD), logged` ~ `GDP / capita (USD), logged`, df_plot)
#summary(mod)
m <- round(mod$coefficients[2], 2)
b <- round(mod$coefficients[1], 2)

# unique(df_plot$Country[which(df_plot$Region == "Asia\n(excluding W. Asia)")])
# unique(df_plot$Country[which(df_plot$Region == "N. Africa / W. Asia")])

df_plot$label_these <- NA
u <- df_plot$Country
df_plot$label_these[which(u %in% labelThese_vec)] <- u[which(u %in% labelThese_vec)]

n <- length(unique(df_plot$Region))
bag_of_colors <- randomcoloR::distinctColorPalette(k = 2 * n)
color_vec <- sample(bag_of_colors, n)

shape_vec <- c(21:24, 4)

gg <- ggplot(df_plot, aes(x = `GDP / capita (USD), logged`,
                          y = `Real expenditures per capita (USD), logged`,
                          group = Region, fill = Region, shape = Region,
                          label = label_these))
# gg <- gg + geom_point(aes(fill = Region, group = Region, shape = Region), pch = 21, alpha = 0.5, size = 6, color = "black")
gg <- gg + geom_point(alpha = 0.6, size = point_size, color = "black", stroke = 1)
gg <- gg + scale_fill_manual(values = color_vec)
gg <- gg + scale_shape_manual(values = shape_vec)
# gg <- gg + geom_abline(slope = m, intercept = b,
#                        linetype = "dashed", size = 1)
gg <- gg + geom_smooth(aes(group = NULL, fill = NULL, shape = NULL), method = lm, se = F)
gg <- gg + geom_text_repel(color = "black", size = label_size)
gg <- gg + labs(#title = "Engel's Law", 
  subtitle = paste0("Slope = ", m, ", y-intercept = ", b))#,
#caption = "Source: Author's creation based on World Bank International Comparison Program Data for 2017.
#GDP/capita is based on 2015 prices.")
gg <- gg + guides(fill = guide_legend(nrow = 2, byrow = T, override.aes = list(linetype = 0)),
                  color = guide_legend(override.aes = list(linetype = 0)))
gg <- gg + theme(legend.position = "bottom",
                 legend.spacing.x = unit(0.5, 'cm'),
                 legend.title = element_blank())
gg
#ggsave("Engels Law 2.png", width = 8.5, height = 6.5, units = "in")

```

Variation 1 disaggregated by region...again note how the fitted parameters vary from one region to another... In this case, Sub-Saharan Africa is more cleanly fit while the other regions are more poorly fit Figure \ref{fig:Engel1_disagg}.

```{r, fig.show = "hold", fig.width = 5, fig.height=6, fig.align='left', fig.cap="\\label{fig:Engel1_disagg}Engel's Law, variation 1, by region, ICP data for 2017.", echo = FALSE}
region_vec <- unique(df_plot$Region)
facet_labels <- c()
for(i in 1:length(region_vec)){
  this_region <- region_vec[i]
  this_df_plot <- subset(df_plot, Region == this_region)
  mod <- lm(`Real expenditures per capita (USD), logged` ~ `GDP / capita (USD), logged`, this_df_plot)
  #summary(mod)
  m <- round(mod$coefficients[2], 2)
  b <- round(mod$coefficients[1], 2)
  #   print(this_region)
  # print(m)
  facet_labels[i] <- paste0("Slope = ", m, ", y-intercept = ", b)
}

names(facet_labels) <- region_vec

gg <- ggplot(df_plot, aes(x = `GDP / capita (USD), logged`,
                          y = `Real expenditures per capita (USD), logged`,
                          group = Region, fill = Region, shape = Region,
                          label = label_these))
gg <- gg + geom_point(alpha = 0.6, size = point_size, color = "black", stroke = 1)
gg <- gg + scale_fill_manual(values = color_vec)
gg <- gg + scale_shape_manual(values = shape_vec)
gg <- gg + geom_smooth(method = lm, se = F)
gg <- gg + facet_wrap(~Region, ncol = 2, scales = "free", labeller = labeller(Region = facet_labels))
gg <- gg + geom_text_repel(color = "black", size = label_size)
gg <- gg + theme(strip.background = element_blank(),
                 strip.text.x = element_text(hjust = -0.01),
                 #legend.direction = "vertical",
                 #legend.spacing.y = unit(1, 'cm'),
                 legend.title = element_blank())
gg <- gg + guides(fill = guide_legend(#nrow = 2,
  override.aes = list(linetype = 0)),
  color = guide_legend(override.aes = list(linetype = 0)))
gg <- shift_legend2(gg)
gg <- ggplotify::as.ggplot(gg)
#gg
#ggsave("Engels Law 2 Disaggregated.png", width = 8.5, height = 6.5, units = "in")
```

Tilman et al. [-@tilman2011global] examined a second variation of Engel's Law and used it to calculate future global food demand.^[Tilman et al. do not mention Engel's Law in their paper, believing this empirical relation to be newly discovered by them.] ...using FAO data...\ref{fig:Engel2}. ...This is also useful since the FAO data extend back to the 1960s, whereas the ICP data extend back only to 2011, and only 2017 for Africa. Therefore it is possible to examine the stability of the empirical relation over a substantial period of time.

```{r, fig.show = "hold", fig.width = 5, fig.height=5, fig.align='left', fig.cap="\\label{fig:Engel2}Engel's Law, variation 2, FAO data for 2017.", echo = FALSE}

df_raw_1 <- read.csv("FoodBalanceSheetsHistoric_E_All_Data.csv", stringsAsFactors = F)
df_raw_1 <- subset(df_raw_1, Item.Code != 2928)
df_raw_2 <- read.csv("FoodBalanceSheets_E_All_Data.csv", stringsAsFactors = F)
df_raw_2 <- subset(df_raw_2, Item.Code != 2928)
rm_cols <- c("Area.Code", "Item.Code", "Element.Code")
rm_cols <- which(colnames(df_raw_1) %in% rm_cols)
df_raw_1 <- df_raw_1[, -rm_cols]
df_raw_2 <- df_raw_2[, -rm_cols]
df_raw <- merge(df_raw_1, df_raw_2, by = c("Area", "Item", "Element", "Unit"))
df_raw$Item <- as.character(df_raw$Item)
df_raw$Element <- as.character(df_raw$Element)
df_raw$Area <- as.character(df_raw$Area)
u <- colnames(df_raw)
df_raw <- df_raw[, -grep("F", u)]
v <- colnames(df_raw)[5:ncol(df_raw)]
colnames(df_raw)[5:ncol(df_raw)] <- gsub("Y", "", v)
df_raw <- gather(df_raw, Year, Value, `1961`:`2017`)
#------------------------------------------------------------------------
rm(df_raw_1, df_raw_2); gc()
#------------------------------------------------------------------------
# out_list <- FAOdat_createRegionGroups(df_raw, exclude_these = NULL, keep_FAOregions = F,
#                                       df_is_trade_matrix = F,
#                                       consolidate_LAC = T,
#                                       consolidate_WEur = T,
#                                       consolidate_AusNZPacIslands = T,
#                                       consolidate_SSA = F,
#                                       output_countryVecs = T)
# df_raw <- out_list[["df_raw"]]
#--------------------------------------------------------------------------
df_raw$Area[grep("Tanzania", df_raw$Area)] <- "Tanzania"
df_raw$Area[grep("C?te d'Ivoire", df_raw$Area)] <- "Cte d'Ivoire"
#--------------------------------------------------------------------------
#unique(df_raw$Element)
bennett_vec <- c("Fat supply quantity (g/capita/day)",
                 "Protein supply quantity (g/capita/day)",
                 "Food supply (kcal/capita/day)")
df_fbalBennett <- subset(df_raw, Element %in% bennett_vec &
                           Item == "Grand Total")
df_fbalBennett <- df_fbalBennett[, c("Area", "Year", "Element", "Value")]
#--------------------------------------------------------------------------
df_fbal <- subset(df_raw, Element == "Food supply (kcal/capita/day)" &
                    Item == "Grand Total")
df_fbal <- df_fbal[, c("Area", "Year", "Value")]
df_fbal$Value <- log(df_fbal$Value)
colnames(df_fbal)[ncol(df_fbal)] <- "Food supply (kcal/capita/day), logged"
colnames(df_fbal)[1] <- "Country"
#rm(df_raw); gc()
#--------------------------------------------------------------------------
df_gdp <- df_gdp <- subset(df_gdp_raw, Item == "Gross Domestic Product per capita" &
                             Element == "Value US$, 2015 prices")
colnames(df_gdp)[1] <- "Country"
df_gdp <- df_gdp[, c("Country", "Year", "Region", "Value")]
df_gdp$Value <- log(df_gdp$Value)
colnames(df_gdp)[ncol(df_gdp)] <- "GDP / capita (USD), logged"
df_engel <- merge(df_fbal, df_gdp, by = c("Country", "Year"))
#--------------------------------------------------------------------------
df_engel$Region[grep("Western Asia", df_engel$Region)] <- "N. Africa / W. Asia"
df_engel$Region[grep("Northern Africa", df_engel$Region)] <- "N. Africa / W. Asia"
df_engel$Region[which(df_engel$Region %in% c("Western Africa",
                                             "Eastern Africa",
                                             "Middle Africa",
                                             "Southern Africa"))] <- "Sub-Saharan Africa"
df_engel$Region[grep("Europe", df_engel$Region)] <- "Europe"
df_engel$Region[which(df_engel$Region %in% c("Europe", "North America"))] <- "Europe / N. Amer.\nAustralia / NZ"
df_engel$Region[which(df_engel$Area %in% c("Australia", "New Zealand"))] <- "Europe / N. Amer.\nAustralia / NZ"
df_engel$Region[which(df_engel$Region %in% c("Southern Asia", "South-Eastern Asia", "Central Asia", "Eastern Asia"))] <- "Asia\n(excluding W. Asia)"
df_engel$Region[which(df_engel$Region == "Aus/NZ/Oceania")] <- "Pacific Islands"
df_engel <- subset(df_engel, Region != "Pacific Islands" &
                     Country != "Bahamas")

#--------------------------------------------------------------------------
#write.csv(df_engel, "Engels Law 3.csv")
this_year <- "2017"
df_plot <- subset(df_engel, Year == this_year)

df_mod <- df_plot[, setdiff(colnames(df_plot), c("Country", "Year", "Region"))]
#keep_rows <- which(!is.na(df_mod$`Consumption Inequality (Gini Coefficient)`))
#df_mod <- df_mod[keep_rows, ]
keep_rows <- which(!is.na(df_mod$`Food supply (kcal/capita/day), logged`))
df_mod <- df_mod[keep_rows, ]

row.names(df_mod) <- df_plot$Country[keep_rows]
#df_mod$`Income Inequality (Gini Coefficient)` <- NULL
#df_mod$`Consumption Inequality (Gini Coefficient)` <- NULL
#df_mod$`Income Inequality (Gini Coefficient)` <- log(df_mod$`Income Inequality (Gini Coefficient)`)
#df_mod$`Consumption Inequality (Gini Coefficient)` <- log(df_mod$`Consumption Inequality (Gini Coefficient)`)

#df_mod$`Consumption Inequality (Gini Coefficient)` <- NULL

mod <- lm(`Food supply (kcal/capita/day), logged` ~ ., df_mod)
#summary(mod)
#m_gini <- round(mod$coefficients[3], 2)
m_gdp <- round(mod$coefficients[2], 2)
b <- round(mod$coefficients[1], 2)

# plot(mod$fitted.values, mod$residuals)
# plot(mod$fitted.values, df_mod$`Food supply (kcal/capita/day), logged`)

# y_lab <- paste(b, "+", m_gdp, "(GDP/capita (USD, 2015 prices), logged) +\n",
#       m_gini, "(Consumption Inequality (Gini Coefficient), logged)")

# df_mod$`6.93 + 0.1 (GDP/capita (USD, 2015 prices), logged) +\n -0.14 (Consumption Inequality (Gini Coefficient), logged)` <-
#   mod$fitted.values
# df_mod$Area <- row.names(df_mod)
# df_mod <- merge(df_mod, df_plot[, c("Area", "Region")], by = "Area")
# df_plot <- df_mod
#unique(df_plot$Area[grep("Korea", df_plot$Area, ignore.case = T)])
df_plot$label_these <- NA
u <- which(df_plot$Country %in% labelThese_vec)
df_plot$label_these[u] <- df_plot$Country[u]

# n <- length(unique(df_plot$Region))
# bag_of_colors <- randomcoloR::distinctColorPalette(k = 2 * n)
# color_vec <- sample(bag_of_colors, n)
gg <- ggplot(df_plot, aes(x = `GDP / capita (USD), logged`,
                          y = `Food supply (kcal/capita/day), logged`,
                          group = Region, fill = Region, shape = Region,
                          label = label_these))
# gg <- gg + geom_point(aes(fill = Region, group = Region, shape = Region), pch = 21, alpha = 0.5, size = 6, color = "black")
gg <- gg + geom_point(alpha = 0.6, size = point_size, color = "black", stroke = 1)
gg <- gg + scale_fill_manual(values = color_vec)
gg <- gg + scale_shape_manual(values = shape_vec)
# gg <- gg + geom_abline(slope = m, intercept = b,
#                        linetype = "dashed", size = 1)
gg <- gg + geom_smooth(aes(group = NULL, fill = NULL, shape = NULL), method = lm, se = F)
gg <- gg + geom_text_repel(color = "black", size = label_size)
gg <- gg + labs(#title = "Engel's Law", 
  subtitle = paste0("Slope = ", m_gdp, ", y-intercept = ", b))#,
#caption = "Source: Author's creation based on World Bank International Comparison Program Data for 2017.
#GDP/capita is based on 2015 prices.")
gg <- gg + guides(fill = guide_legend(nrow = 2, byrow = T, override.aes = list(linetype = 0)),
                  color = guide_legend(override.aes = list(linetype = 0)))
gg <- gg + theme(legend.position = "bottom",
                 legend.spacing.x = unit(0.5, 'cm'),
                 legend.title = element_blank())
gg
#ggsave("Engels Law 3.png", width = 8.5, height = 6.5, units = "in")


```

Perhaps the main benefit of this variation on Engel's Law is that the data go back a long ways, so you can investigate its stability over time. The model Figure \ref{fig:Engel_stability}. An animation is also provided in the supplemental materials.

```{r, fig.show = "hold", fig.width = 4, fig.height=3, fig.align='left', fig.cap="\\label{fig:Engel_stability}Engel's Law parameter stability, FAO data 1970-2018.", echo = FALSE}

df_params <- df_engel[, c("Year", "Food supply (kcal/capita/day), logged",
                          "GDP / capita (USD), logged")]
df_params <- df_params[which(!is.na(df_params$`Food supply (kcal/capita/day), logged`)), ]

#write.csv(df_params, "Engels Law stability.csv")

yr_vec <- unique(df_params$Year)
n_yrs <- length(yr_vec)
m_vec <- c()
b_vec <- c()
adjR2_vec <- c()
for(i in 1:n_yrs){
  this_year <- yr_vec[i]
  this_df <- subset(df_params, Year == this_year)
  this_df$Year <- NULL
  mod <- lm(`Food supply (kcal/capita/day), logged`~., this_df)
  m_vec[i] <- round(mod$coefficients[2], 3)
  b_vec[i] <- round(mod$coefficients[1], 3)
  df_out <- as.data.frame(broom::glance(mod))
  adjR2_vec[i] <- df_out$adj.r.squared
  
}

df_plot <- data.frame(Year = yr_vec, Slope = m_vec, b_vec, adjR2_vec)
colnames(df_plot)[3:4] <- c("Y-intercept", "Adj. R squared")
mean(df_plot$Slope)
sd(df_plot$Slope)
mean(df_plot$`Y-intercept`)
sd(df_plot$`Y-intercept`)

df_plot$Year <- as.integer(as.character(df_plot$Year))
df_plot <- df_plot %>% gather(Type, Value, Slope:`Adj. R squared`)
gg <- ggplot(df_plot, aes(x = Year, y = Value))
gg <- gg + geom_line()
gg <- gg + facet_wrap(~Type, ncol = 1, scales = "free_y")
gg <- gg + theme(axis.title = element_blank(),
                 strip.background = element_blank())
gg


```

An important implication of Engel's Law is that distribution matters for food demand. Adding consumption inequality to the equation increases the fit...

```{r, results="asis", message=FALSE, echo = FALSE}

df_CIneq_raw <- read.csv("consumption-inequalityexcel.csv", stringsAsFactors = F)
# colnames(df_CIneq_raw)
# unique(df_CIneq_raw$country)
# unique(df_CIneq_raw$year)
df_CIneq_raw$year <- as.integer(df_CIneq_raw$year)
df_CIneq_raw <- df_CIneq_raw[, c("country", "year", "gini")]
df_CIneq_raw$Type <- "Consumption Inequality (Gini Coefficient)"
df_Ineq <- df_CIneq_raw
colnames(df_Ineq)[1:2] <- c("Area", "Year")
#==========================================================================
df_Ineq$Area[grep("Tanzania", df_Ineq$Area, ignore.case = T)] <- "Tanzania"
df_Ineq$Area[grep("Iran", df_Ineq$Area, ignore.case = T)] <- "Iran"
df_Ineq$Area[grep("Syria", df_Ineq$Area, ignore.case = T)] <- "Syria"
df_Ineq$Area[grep("United States Virgin Islands", df_Ineq$Area, ignore.case = T)] <- "US Virgin Islands"
df_Ineq$Area[grep("British Virgin Islands", df_Ineq$Area, ignore.case = T)] <- "UK Virgin Islands"
df_Ineq$Area[grep("United States", df_Ineq$Area, ignore.case = T)] <- "United States"
df_Ineq$Area[grep("Viet", df_Ineq$Area, ignore.case = T)] <- "Vietnam"
df_Ineq$Area[grep("Democratic Republic of the Congo", df_Ineq$Area, ignore.case = T)] <- "DRC"
df_Ineq$Area[grep("Congo, Dem. Rep.", df_Ineq$Area, ignore.case = T)] <- "DRC"
df_Ineq$Area[grep("Bolivia", df_Ineq$Area, ignore.case = T)] <- "Bolivia"
df_Ineq$Area[grep("Singapore", df_Ineq$Area, ignore.case = T)] <- "Singapore"
df_Ineq$Area[grep("Macedonia", df_Ineq$Area, ignore.case = T)] <- "Macedonia"
df_Ineq$Area[grep("Venezuela", df_Ineq$Area, ignore.case = T)] <- "Venezuela"
df_Ineq$Area[grep("Ivoire", df_Ineq$Area, ignore.case = T)] <- "Cote d'Ivoire"
df_Ineq$Area[grep("Lanka", df_Ineq$Area, ignore.case = T)] <- "Sri Lanka"
df_Ineq$Area[grep("Swaziland", df_Ineq$Area, ignore.case = T)] <- "Eswatini"
df_Ineq$Area[grep("Lao", df_Ineq$Area, ignore.case = T)] <- "Lao, PDR"

#    unique(df_Ineq$Area[grep("Lao", df_Ineq$Area, ignore.case = T)])



colnames(df_Ineq)[1] <- "Country"
colnames(df_Ineq)[3] <- "Consumption Inequality (Gini Coefficient)"

#==========================================================================
this_year <- "2012"
df_mod1 <- subset(df_engel, Year == this_year)
df_mod2 <- subset(df_Ineq, Year == this_year)
# dont_match1 <- setdiff(unique(df_mod1$Country), unique(df_mod2$Country))
# dont_match2 <- setdiff(unique(df_mod2$Country), unique(df_mod1$Country))
#     unique(df_Ineq$Country[grep("United", df_Ineq$Country, ignore.case = T)])
# unique(df_engel$Country[grep("United", df_engel$Country, ignore.case = T)])
df_mod <- merge(df_mod1, df_mod2, by = c("Country"))
df_mod <- df_mod[, c("Country", "Food supply (kcal/capita/day), logged",
                     "GDP / capita (USD), logged",
                     "Consumption Inequality (Gini Coefficient)")]

#write.csv(df_mod, "Engels Law regression.csv")

df_mod$Country <- NULL
df_mod$`Consumption Inequality (Gini Coefficient)` <- log(df_mod$`Consumption Inequality (Gini Coefficient)`)
colnames(df_mod)[ncol(df_mod)] <- "Consumption Inequality (Gini Coefficient), logged"
#df_mod$`Consumption Inequality (Gini Coefficient), logged` <- NULL

mod <- lm(`Food supply (kcal/capita/day), logged` ~ ., df_mod)
#summary(mod)

stargazer::stargazer(mod, type = "latex",
                     title = "Including consumption inequality in Engel's Law improves the fit",
                     header = FALSE,
                     single.row = TRUE)

```

## Bennett's Law

4 calories in a gram of protein or carbohydrate, 9 calories in a gram of fat.
https://www.nal.usda.gov/fnic/how-many-calories-are-one-gram-fat-carbohydrate-or-protein

...Figure \ref{fig:Bennett0}. This means that the protein share of the diet remains fairly constant, although protein sources may vary--passing from legumes to animal proteins, for example. The correlation with higher income can also be viewed in this variation by the size of the points.

```{r, fig.show = "hold", fig.width = 5, fig.height=5, fig.align='left', fig.cap="\\label{fig:Bennett0}Bennett's Law, FAO data for 2017.", echo = FALSE}

df_bennett <- df_fbalBennett
#colnames(df_bennett)
df_bennett <- df_bennett %>% spread(Element, Value)
df_bennett$`Fat supply quantity (kcal/capita/day)` <- 9 * df_bennett$`Fat supply quantity (g/capita/day)`
df_bennett$`Protein supply quantity (kcal/capita/day)` <- 4 * df_bennett$`Protein supply quantity (g/capita/day)`

df_bennett$`Fat supply quantity (g/capita/day)` <- NULL
df_bennett$`Protein supply quantity (g/capita/day)` <- NULL

df_bennett$`Carb supply quantity (kcal/capita/day)` <-
  df_bennett$`Food supply (kcal/capita/day)` - 
  df_bennett$`Fat supply quantity (kcal/capita/day)` -
  df_bennett$`Protein supply quantity (kcal/capita/day)`

df_bennett$`Fat share of diet (%)` <- 100 * df_bennett$`Fat supply quantity (kcal/capita/day)` / df_bennett$`Food supply (kcal/capita/day)`
df_bennett$`Carb share of diet (%)` <- 100 * df_bennett$`Carb supply quantity (kcal/capita/day)` / df_bennett$`Food supply (kcal/capita/day)`

df_bennett$`Carb share of diet (%), logged` <- log(df_bennett$`Carb share of diet (%)`)

colnames(df_bennett)[1] <- "Country"
df_bennett <- merge(df_bennett, df_gdp, by = c("Country", "Year"))
#---------------------------------------------------------------------------
df_bennett$Region[grep("Western Asia", df_bennett$Region)] <- "N. Africa / W. Asia"
df_bennett$Region[grep("Northern Africa", df_bennett$Region)] <- "N. Africa / W. Asia"
df_bennett$Region[which(df_bennett$Region %in% c("Western Africa",
                                                 "Eastern Africa",
                                                 "Middle Africa",
                                                 "Southern Africa"))] <- "Sub-Saharan Africa"
df_bennett$Region[grep("Europe", df_bennett$Region)] <- "Europe"
df_bennett$Region[which(df_bennett$Region %in% c("Europe", "North America"))] <- "Europe / N. Amer.\nAustralia / NZ"
df_bennett$Region[which(df_bennett$Area %in% c("Australia", "New Zealand"))] <- "Europe / N. Amer.\nAustralia / NZ"
df_bennett$Region[which(df_bennett$Region %in% c("Southern Asia", "South-Eastern Asia", "Central Asia", "Eastern Asia"))] <- "Asia\n(excluding W. Asia)"
df_bennett$Region[which(df_bennett$Region == "Aus/NZ/Oceania")] <- "Pacific Islands"
df_bennett <- subset(df_bennett, Region != "Pacific Islands")
#unique(df_bennett$Region)
#------------------------------------------------------------------------
#write.csv(df_bennett, "Bennetts Law data.csv")
df_plot <- subset(df_bennett, Year == "2017")
df_mod <- df_plot[, c("Carb share of diet (%), logged", "GDP / capita (USD), logged")]
mod <- lm(`Carb share of diet (%), logged` ~., data = df_mod)
#summary(mod)
m <- round(mod$coefficients[2], 2)
b <- round(mod$coefficients[1], 2)

df_plot$label_these <- NA
u <- df_plot$Country
df_plot$label_these[which(u %in% labelThese_vec)] <- df_plot$Country[which(u %in% labelThese_vec)]

gg <- ggplot(df_plot, aes(x = `GDP / capita (USD), logged`,
                          y = `Carb share of diet (%), logged`,             
                          group = Region, fill = Region,
                          shape = Region,
                          label = label_these))
# gg <- gg + geom_point(aes(fill = Region, group = Region, shape = Region), pch = 21, alpha = 0.5, size = 6, color = "black")
gg <- gg + geom_point(alpha = 0.6, size = point_size, color = "black", stroke = 1)
gg <- gg + scale_fill_manual(values = color_vec)
gg <- gg + scale_shape_manual(values = shape_vec)
# gg <- gg + geom_abline(slope = m, intercept = b,
#                        linetype = "dashed", size = 1)
gg <- gg + geom_smooth(aes(group = NULL, fill = NULL, shape = NULL), method = lm, se = F)
gg <- gg + geom_text_repel(color = "black", size = label_size)
gg <- gg + labs(#title = "Engel's Law", 
  subtitle = paste0("Slope = ", m, ", y-intercept = ", b))#,
#caption = "Source: Author's creation based on World Bank International Comparison Program Data for 2017.
#GDP/capita is based on 2015 prices.")
gg <- gg + guides(fill = guide_legend(nrow = 2, byrow = T, override.aes = list(linetype = 0)),
                  color = guide_legend(override.aes = list(linetype = 0)))
gg <- gg + theme(legend.position = "bottom",
                  legend.spacing.x = unit(0.5, 'cm'),
                  legend.title = element_blank())
gg

```

...Figure \ref{fig:Bennett1}

```{r, fig.show = "hold", fig.width = 5, fig.height=5, fig.align='left', fig.cap="\\label{fig:Bennett1}A variation of Bennett's Law, FAO data for 2017.", echo = FALSE}

#------------------------------------------------------------------------

df_mod <- df_plot[, c("Carb share of diet (%)", "Fat share of diet (%)")]
mod <- lm(`Carb share of diet (%)` ~., data = df_mod)
#summary(mod)
m <- round(mod$coefficients[2], 2)
b <- round(mod$coefficients[1], 2)

gg <- ggplot(df_plot, aes(x = `Fat share of diet (%)`,
                          y = `Carb share of diet (%)`,         
                          group = Region, fill = Region,
                          shape = Region,
                          label = label_these))
# gg <- gg + geom_point(aes(fill = Region, group = Region, shape = Region), pch = 21, alpha = 0.5, size = 6, color = "black")
gg <- gg + geom_point(aes(size = `GDP / capita (USD), logged`), alpha = 0.6, color = "black", stroke = 1)
gg <- gg + scale_fill_manual(values = color_vec)
gg <- gg + scale_shape_manual(values = shape_vec)
# gg <- gg + geom_abline(slope = m, intercept = b,
#                        linetype = "dashed", size = 1)
gg <- gg + geom_smooth(aes(group = NULL, fill = NULL, shape = NULL), method = lm, se = F)
gg <- gg + geom_text_repel(color = "black", size = label_size)
gg <- gg + labs(#title = "Engel's Law", 
  subtitle = paste0("Slope = ", m, ", y-intercept = ", b))#,
#caption = "Source: Author's creation based on World Bank International Comparison Program Data for 2017.
#GDP/capita is based on 2015 prices.")
gg <- gg + guides(fill = guide_legend(nrow = 2, byrow = T, override.aes = list(linetype = 0)),
                  color = guide_legend(override.aes = list(linetype = 0)))
gg <- gg + theme(legend.position = "bottom",
                  legend.spacing.x = unit(0.5, 'cm'),
                  legend.title = element_blank())
gg
# gg1 + gg2 + plot_layout(ncol = 2, guides = "collect") & theme(legend.position = 'bottom')


```












```{r, fig.show = "hold", fig.width = 5, fig.height=4, fig.align='left', fig.cap="\\label{fig:price_series}The higher the price, the more volatile. FAO data 1972-2017.", echo = FALSE}

df_export_raw <- read.csv("Trade_Crops_Livestock_E_All_Data.csv", stringsAsFactors = F)
df_export_raw <- subset(df_export_raw, Item.Code != 2928)
df_export_raw$Area.Code <- NULL
df_export_raw$Element.Code <- NULL
df_export_raw$Item <- as.character(df_export_raw$Item)
df_export_raw$Element <- as.character(df_export_raw$Element)
df_export_raw$Area <- as.character(df_export_raw$Area)
u <- colnames(df_export_raw)
df_export_raw <- df_export_raw[, -grep("F", u)]
colnames(df_export_raw)[6:ncol(df_export_raw)] <- as.character(c(1961:2017))
df_export_raw <- gather(df_export_raw,Year,Value,`1961`:`2017`)
rm(u)
df_export_raw$Year <- as.integer(df_export_raw$Year)
#---------------------------------------------------------------------------
#unique(df_export_raw$Area)[grep("World", unique(df_export_raw$Area))]
element_vec <- c("Export Quantity", "Export Value")
df_export_world <- subset(df_export_raw[, c("Area", "Year", "Element", "Item", "Value")], Area == "World" & Element %in% element_vec)
df_export_world <- df_export_world %>% spread(Element, Value)
#---------------------------------------------------------------------------
df_export_world <- df_export_world[which(!is.na(df_export_world$`Export Quantity`)), ]
df_export_world <- df_export_world[which(!is.na(df_export_world$`Export Value`)), ]
df_export_world <- df_export_world[which(df_export_world$`Export Quantity` != 0), ]

df_export_world$`World Export Price (USD / metric ton)` <- NA
df_export_world$`World Export Price (USD / metric ton)` <- 1000 * as.numeric(df_export_world$`Export Value`) / as.numeric(df_export_world$`Export Quantity`)
#df_export_world$`Export Quantity` <- NULL
#df_export_world$`Export Value` <- NULL

#unique(df_export_world$Item)
plot_these <- c("Maize", "Apples", "Cocoa, beans", "Cocoa, butter")
df_plot <- subset(df_export_world, Item %in% plot_these &
                    Year >= 1972)
yr_1 <- unique(df_plot$Year)[1]
yr_n <- max(unique(df_plot$Year))
gg <- ggplot(df_plot, aes(x = Year, y = `World Export Price (USD / metric ton)`, group = Item, color = Item)) + geom_line(lwd = 1)
gg <- gg + scale_x_continuous(breaks = seq(yr_1, yr_n, length.out = 6))
#gg <- gg + scale_y_log10()
gg <- gg + theme(axis.title.x = element_blank(),
                 legend.title = element_blank(),
                 legend.position = "top")
gg

```


```{r, fig.show = "hold", fig.width = 5, fig.height=4, fig.align='left', fig.cap="\\label{fig:risk_reward}Logged expected price plotted against logged standard deviation. FAO data 2016-2017.", echo = FALSE}

df_riskReward <- df_export_world %>% group_by(Item) %>%
  summarise(`Mean World Export Price (USD / metric ton)` = mean(`World Export Price (USD / metric ton)`, na.rm = T),
            `Stand. Dev. World Export Price (USD / metric ton)` = sd(`World Export Price (USD / metric ton)`, na.rm = T),
            `Mean Export Qty.` = mean(`Export Quantity`, na.rm = T), 
            `Mean Export Value ('000 USD)` = mean(`Export Value`, na.rm = T))

df_riskReward$`Mean World Export Price (USD / metric ton), logged` <-
  log(df_riskReward$`Mean World Export Price (USD / metric ton)`)
df_riskReward$`Stand. Dev. World Export Price (USD / metric ton), logged` <-
  log(df_riskReward$`Stand. Dev. World Export Price (USD / metric ton)`)
#--
df_riskReward <- df_riskReward[which(!is.na(df_riskReward$`Stand. Dev. World Export Price (USD / metric ton), logged`)), ]
#nrow(df_riskReward)
#----------------------------------------------------------------------------
df_mod <- df_riskReward[, c("Mean World Export Price (USD / metric ton), logged", "Stand. Dev. World Export Price (USD / metric ton), logged")]
#nrow(df_mod)
row.names(df_mod) <- df_riskReward$Item
mod <- lm(`Mean World Export Price (USD / metric ton), logged` ~., df_mod)
#summary(mod)
b <- round(mod$coefficients[1], 2)
m <- round(mod$coefficients[2], 2)
#---
gg <- ggplot(data.frame(Item = row.names(df_mod), yhat = mod$fitted.values,
                        resid = mod$residuals),
             aes(x = yhat, y = resid, label = Item))
gg <- gg + geom_point() + geom_text()
gg <- gg + geom_hline(yintercept = c(-1, 1), color = "red")
gg
#---
ind_outlier <- which(abs(as.numeric(mod$residuals)) >= 0.75)
excluded_items <- row.names(df_mod)[ind_outlier]
pct_removed <- 100 * length(ind_outlier) / nrow(df_riskReward)
row_names <- row.names(df_mod)[-ind_outlier]
df_mod <- df_mod[-ind_outlier, ]
row.names(df_mod) <- row_names
mod <- lm(`Mean World Export Price (USD / metric ton), logged` ~., df_mod)
#summary(mod)
b <- round(mod$coefficients[1], 2)
m <- round(mod$coefficients[2], 2)
#plot(mod$fitted.values, mod$residuals)
#----------------------------------------------------------------------------
#df_riskReward$Item[ind_outlier]
df_plot <- df_riskReward[-ind_outlier, c("Item", "Mean World Export Price (USD / metric ton), logged", "Stand. Dev. World Export Price (USD / metric ton), logged",
                             "Mean Export Value ('000 USD)")]

df_plot$`Mean Export Value\n(million USD)` <- df_plot$`Mean Export Value ('000 USD)` / 1000

labelItems <- c("Maize", "Cassava dried", "Flour, wheat", "Starch, cassava",
              "Coffee, green", "Coffee, roasted", "Wheat", "Bananas",
              "Plantains", "Potatoes", "Cassava Equivalent", "Cocoa, beans",
              "Wool, greasy", "Beans, dry", "Cotton lint", "Silk", "Apples",
              "Soybeans", "Oil, soybean", "Rubber, natural", "Rice",
              "Cocoa, butter")

df_plot$label_these <- NA
u <- df_plot$Item
df_plot$label_these[which(u %in% labelItems)] <- u[which(u %in% labelItems)]

label_size <- 5
gg <- ggplot(df_plot, aes(x = `Stand. Dev. World Export Price (USD / metric ton), logged`,
                          y = `Mean World Export Price (USD / metric ton), logged`))
gg <- gg + geom_smooth(method = lm, se = F)
gg <- gg + geom_point(aes(size = `Mean Export Value\n(million USD)`),
                      fill = "khaki1", pch=21, color="Black")
gg <- gg + geom_text_repel(aes(label = label_these), size = label_size)
gg <- gg + labs(title = "The risk-return frontier", subtitle = paste0("Slope = ", m, "  Y intercept = ", b))
gg <- gg + theme(legend.position = "bottom")
gg <- gg + labs(caption = "Source: Author's creation using FAOSTAT export value and quantity series 1961-2017.")
gg
#ggsave("risk reward frontier.pdf")
```

```{r, fig.show = "hold", fig.width = 5, fig.height=4, fig.align='left', fig.cap="\\label{fig:risk_rewardDisagg}Logged expected price plotted against logged standard deviation. FAO data for 1961-2017.", echo = FALSE}
out_list <- FAOdat_createRegionGroups(df_export_raw, exclude_these = NULL,
                                      keep_FAOregions = F,
                                      df_is_trade_matrix = F,
                                      consolidate_LAC = T,
                                      consolidate_WEur = T,
                                      consolidate_AusNZPacIslands = T,
                                      consolidate_SSA = F,
                                      output_countryVecs = T)
df_export_regions <- out_list[["df_raw"]]
# #---------------------------------------------------------------------------
#unique(df_export_raw$Item)
# which(unique(df_export_raw$Item) == "Rice")
# unique(df_export_raw$Item)[105]
#unique(df_export_raw$Item)[grep("Rubber",unique(df_export_raw$Item))]
df_export_regions <- subset(df_export_regions[, c("Area", "Region", "Year", "Element", "Item", "Value")], Element %in% element_vec)
#colnames(df_export_regions)

df_export_regions <- df_export_regions %>% spread(Element, Value)

df_export_regions$Area <- NULL
df_export_regions <- df_export_regions %>% 
  group_by(Region, Year, Item) %>% summarise_all(sum, na.rm = T)
#unique(df_export_regionss$Region)
#--------------------------------------------------------------------------
df_export_regions$Region[grep("Western Asia", df_export_regions$Region)] <- "N. Africa / W. Asia"
df_export_regions$Region[grep("Northern Africa", df_export_regions$Region)] <- "N. Africa / W. Asia"
df_export_regions$Region[which(df_export_regions$Region %in% c("Western Africa",
                                             "Eastern Africa",
                                             "Middle Africa",
                                             "Southern Africa"))] <- "Sub-Saharan Africa"
df_export_regions$Region[grep("Europe", df_export_regions$Region)] <- "Europe"
df_export_regions$Region[which(df_export_regions$Region %in% c("Europe", "North America"))] <- "Europe / N. Amer.\nAustralia / NZ"
df_export_regions$Region[which(df_export_regions$Area %in% c("Australia", "New Zealand"))] <- "Europe / N. Amer.\nAustralia / NZ"
df_export_regions$Region[which(df_export_regions$Region %in% c("Southern Asia", "South-Eastern Asia", "Central Asia", "Eastern Asia"))] <- "Asia\n(excluding W. Asia)"
df_export_regions$Region[which(df_export_regions$Region == "Aus/NZ/Oceania")] <- "Pacific Islands"
df_export_regions <- subset(df_export_regions, Region != "Pacific Islands")
#--------------------------------------------------------------------------
df_export_regions <- df_export_regions[which(!is.na(df_export_regions$`Export Quantity`)), ]
df_export_regions <- df_export_regions[which(!is.na(df_export_regions$`Export Value`)), ]
df_export_regions <- df_export_regions[which(df_export_regions$`Export Quantity` != 0), ]
df_export_regions$`World Export Price (USD / metric ton)` <- NA
df_export_regions$`World Export Price (USD / metric ton)` <- 1000 * as.numeric(df_export_regions$`Export Value`) / as.numeric(df_export_regions$`Export Quantity`)
#write.csv(df_export_regions, "FAO export by regions.csv")
#--------------------------------------------------------------------------
#colnames(df_export_regions)
df_riskReward_regions <- df_export_regions %>% group_by(Region, Item) %>%
  summarise(`Mean World Export Price (USD / metric ton)` = mean(`World Export Price (USD / metric ton)`, na.rm = T),
            `Stand. Dev. World Export Price (USD / metric ton)` = sd(`World Export Price (USD / metric ton)`, na.rm = T),
            `Mean Export Qty.` = mean(`Export Quantity`, na.rm = T), 
            `Mean Export Value ('000 USD)` = mean(`Export Value`, na.rm = T))

df_riskReward_regions$`Mean World Export Price (USD / metric ton), logged` <-
  log(df_riskReward_regions$`Mean World Export Price (USD / metric ton)`)
df_riskReward_regions$`Stand. Dev. World Export Price (USD / metric ton), logged` <-
  log(df_riskReward_regions$`Stand. Dev. World Export Price (USD / metric ton)`)
#--
df_riskReward_regions <- df_riskReward_regions[which(!is.na(df_riskReward_regions$`Stand. Dev. World Export Price (USD / metric ton), logged`)), ]
ind_rm <- which(is.infinite(df_riskReward_regions$`Stand. Dev. World Export Price (USD / metric ton), logged`))
df_riskReward_regions$Item[ind_rm]
df_riskReward_regions <- df_riskReward_regions[-ind_rm, ]
#nrow(df_riskReward_regions)
#----------------------------------------------------------------------------
gg <- ggplot(df_plot, aes(x = `Stand. Dev. World Export Price (USD / metric ton), logged`,
                          y = `Mean World Export Price (USD / metric ton), logged`))
#gg <- gg + geom_smooth(method = lm, se = F)
gg <- gg + geom_point(aes(size = `Mean Export Value\n(million USD)`),
                      fill = "khaki1", pch=21, color="Black")
#gg <- gg + geom_text_repel(aes(label = label_these), size = label_size)
gg <- gg + facet_wrap(~Region)
#gg <- gg + labs(subtitle = paste0("Slope = ", m_vec[i], "  Y intercept = ", b_vec[i]))
# gg <- gg + labs(caption = "Source: Author's creation using FAOSTAT export value and quantity series 1961-2017.")
gg <- shift_legend2(gg)
gg <- ggplotify::as.ggplot(gg)


# n_regions <- length(unique(df_riskReward_regions$Region))
# region_vec <- unique(df_riskReward_regions$Region)
# facet_labels <- c()
# for(i in 1:n_regions){
#   this_region <- region_vec[i]
#   this_df <- subset(df_riskReward_regions, Region == this_region &
#                       !(Item %in% excluded_items))
#   df_mod <- this_df[, c("Mean World Export Price (USD / metric ton), logged", "Stand. Dev. World Export Price (USD / metric ton), logged")]
#   ind <- which(is.infinite(df_mod$`Stand. Dev. World Export Price (USD / metric ton), logged`))
#   this_df$Item[ind]
#   #nrow(df_mod)
#   mod <- lm(`Mean World Export Price (USD / metric ton), logged` ~., df_mod)
#   #summary(mod)
#   b <- round(mod$coefficients[1], 2)
#   m <- round(mod$coefficients[2], 2)
#   #   #---
#   # gg <- ggplot(data.frame(Item = row.names(df_mod), yhat = mod$fitted.values,
#   #                         resid = mod$residuals),
#   #              aes(x = yhat, y = resid, label = Item))
#   # gg <- gg + geom_point() + geom_text()
#   # gg <- gg + geom_hline(yintercept = c(-1, 1), color = "red")
#   # gg
#   # #---
#   ind_outlier <- which(abs(as.numeric(mod$residuals)) >= 0.75)
#   # excluded_items <- row.names(df_mod)[ind_outlier]
#   # pct_removed <- 100 * length(ind_outlier) / nrow(df_riskReward_regions)
#   row_names <- row.names(df_mod)[-ind_outlier]
#   df_mod <- df_mod[-ind_outlier, ]
#   row.names(df_mod) <- row_names
#   mod <- lm(`Mean World Export Price (USD / metric ton), logged` ~., df_mod)
#   #summary(mod)
#   b <- round(mod$coefficients[1], 2)
#   m <- round(mod$coefficients[2], 2)
#   # #plot(mod$fitted.values, mod$residuals)
#   facet_labels[i] <- paste0("Slope = ", m, ", y-intercept = ", b)
#   
#   
# }
# #----------------------------------------------------------------------------
# df_plot <- df_riskReward_regions
# 
# df_plot$`Mean Export Value\n(million USD)` <- df_plot$`Mean Export Value ('000 USD)` / 1000
# 
# df_plot$label_these <- NA
# u <- df_plot$Item
# df_plot$label_these[which(u %in% labelItems)] <- u[which(u %in% labelItems)]
# 
# gg <- ggplot(df_plot, aes(x = `Stand. Dev. World Export Price (USD / metric ton), logged`,
#                           y = `Mean World Export Price (USD / metric ton), logged`,
#                           size = `Mean Export Value\n(million USD)`,
#                           #                          group = Region, fill = Region, shape = Region,
#                           label = label_these))
# gg <- gg + geom_point(alpha = 0.6, color = "black", stroke = 1)
# # gg <- gg + scale_fill_manual(values = color_vec)
# # gg <- gg + scale_shape_manual(values = shape_vec)
# gg <- gg + geom_smooth(method = lm, se = F)
# gg <- gg + facet_wrap(~Region, ncol = 3, scales = "free", labeller = labeller(Region = facet_labels))
# gg <- gg + geom_text_repel(color = "black", size = label_size)
# gg <- gg + theme(strip.background = element_blank(),
#                  strip.text.x = element_text(hjust = -0.01),
#                  #legend.direction = "vertical",
#                  #legend.spacing.y = unit(1, 'cm'),
#                  legend.title = element_blank())
# # gg <- gg + guides(fill = guide_legend(#nrow = 2,
# #   override.aes = list(linetype = 0)),
# #   color = guide_legend(override.aes = list(linetype = 0)))
# gg <- shift_legend2(gg)
# gg <- ggplotify::as.ggplot(gg)
```

# Discussion and conclusion

Policy implications etc. higher expected return goes hand in hand with higher risk---in an almost perfect one-to-one tradeoff. Focus is always on productivity, genetic gain...with little or no attention paid to the role of price risk. but this shows that innovative, cooperative risk pooling mechanisms may be an equally important, complementary role. social capital to mitigate risk...resilience...agricultural innovation systems (Belo Horizonte, Multi-Stake... Platforms?)

# References {-}

<div id="refs"></div>

# Appendix {-}

...Figure \ref{fig:Bennett0_stability}

```{r, fig.show = "hold", fig.width = 5, fig.height=4, fig.align='left', fig.cap="\\label{fig:Bennett0_stability}Bennett's Law stability over time. \\textit{(Right)}A cleaner fitting variation of Bennett's Law. FAO data for 2017.", echo = FALSE}

df_params <- df_bennett[, c("Year", "Carb share of diet (%), logged",
                            "GDP / capita (USD), logged")]
df_params <- df_params[which(!is.na(df_params$`Carb share of diet (%), logged`)), ]

yr_vec <- unique(df_params$Year)
n_yrs <- length(yr_vec)
m_vec <- c()
b_vec <- c()
adjR2_vec <- c()
for(i in 1:n_yrs){
  this_year <- yr_vec[i]
  this_df <- subset(df_params, Year == this_year)
  this_df$Year <- NULL
  mod <- lm(`Carb share of diet (%), logged`~., this_df)
  m_vec[i] <- round(mod$coefficients[2], 3)
  b_vec[i] <- round(mod$coefficients[1], 3)
  df_out <- as.data.frame(broom::glance(mod))
  adjR2_vec[i] <- df_out$adj.r.squared
  
}

df_plot <- data.frame(Year = yr_vec, Slope = m_vec, b_vec, adjR2_vec)
colnames(df_plot)[3:4] <- c("Y-intercept", "Adj. R squared")
mean(df_plot$Slope)
sd(df_plot$Slope)
mean(df_plot$`Y-intercept`)
sd(df_plot$`Y-intercept`)

df_plot$Year <- as.integer(as.character(df_plot$Year))
df_plot <- df_plot %>% gather(Type, Value, Slope:`Adj. R squared`)
gg <- ggplot(df_plot, aes(x = Year, y = Value))
gg <- gg + geom_line()
gg <- gg + scale_x_continuous(breaks = seq(yr_vec[1], yr_vec[n_yrs - 1], length.out = 24))
gg <- gg + facet_wrap(~Type, ncol = 1, scales = "free_y")
gg1 <- gg + theme(axis.title = element_blank(),
                  axis.text.x = element_text(angle = 60, hjust = 1),
                  strip.background = element_blank())



df_params <- df_bennett[, c("Year", "Carb share of diet (%)",
                            "Fat share of diet (%)")]
df_params <- df_params[which(!is.na(df_params$`Carb share of diet (%)`)), ]

yr_vec <- unique(df_params$Year)
n_yrs <- length(yr_vec)
m_vec <- c()
b_vec <- c()
adjR2_vec <- c()
for(i in 1:n_yrs){
  this_year <- yr_vec[i]
  this_df <- subset(df_params, Year == this_year)
  this_df$Year <- NULL
  mod <- lm(`Carb share of diet (%)`~., this_df)
  m_vec[i] <- round(mod$coefficients[2], 3)
  b_vec[i] <- round(mod$coefficients[1], 3)
  df_out <- as.data.frame(broom::glance(mod))
  adjR2_vec[i] <- df_out$adj.r.squared
  
}

df_plot <- data.frame(Year = yr_vec, Slope = m_vec, b_vec, adjR2_vec)
colnames(df_plot)[3:4] <- c("Y-intercept", "Adj. R squared")
mean(df_plot$Slope)
sd(df_plot$Slope)
mean(df_plot$`Y-intercept`)
sd(df_plot$`Y-intercept`)

df_plot$Year <- as.integer(as.character(df_plot$Year))
df_plot <- df_plot %>% gather(Type, Value, Slope:`Adj. R squared`)
gg <- ggplot(df_plot, aes(x = Year, y = Value))
gg <- gg + geom_line()
gg <- gg + scale_x_continuous(breaks = seq(yr_vec[1], yr_vec[n_yrs - 1], length.out = 24))
gg <- gg + facet_wrap(~Type, ncol = 1, scales = "free_y")
gg2 <- gg + theme(axis.title = element_blank(),
                  axis.text.x = element_text(angle = 60, hjust = 1),
                  strip.background = element_blank())

gg1 + gg2 + plot_layout(ncol = 2)

rm(df_raw, df_gdp_raw, df_icp_raw, df_CIneq_raw)

```
